---
title: "Annotate PDZs structure metrics"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---
# libraries used for the analysis
```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(bio3d)
library(car)
library(systemfonts)
library(plyr)
library(ggiraph)
library(ggiraphExtra)
library(rstatix)
library(data.table)
library(dplyr)
library(seqinr)
library(stringr)
library(ggplot2)
library(Matrix)
library(viridis)
library(ggpubr)
library(readxl)
library(stringi)
library(GGally)
library("readxl")
library(modelr)
library(cowplot)
library(ggpointdensity)
library("Biostrings")
library(tidyr)
```



# variables
```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
setwd(base_dir)

# load some data
load(paste0('source_variables.RData'))
assays_info<-as.data.table(read_xlsx("data/assays_info.xlsx"))
assays_info[,library:=paste0(pdz_name, "_", ligand_name)]

aa_mapping <- c(
  ALA = "A", ARG = "R", ASN = "N", ASP = "D", CYS = "C", 
  GLU = "E", GLN = "Q", GLY = "G", HIS = "H", ILE = "I", 
  LEU = "L", LYS = "K", MET = "M", PHE = "F", PRO = "P", 
  SER = "S", THR = "T", TRP = "W", TYR = "Y", VAL = "V"
)
```


# 2. SECONDARY STRUCTURE INFO (helix,sheet,rsasa)
```{r}
# Initialize data.table to store secondary structures
secondary_structures <- data.table()

# Process each PDZ library
for (pdz_name in libraries_folding_names) {
  print(paste0("reading: data/structures/alphafold3/", pdz_name, ".pdb"))
  
  # Read PDB structure
  struct <- read.pdb(paste0("data/structures/alphafold3/", pdz_name, ".pdb"))
  pdb_table <- as.data.table(struct[1])
  
  # Extract chain A sequence and residue numbers
  residue_numbers <- sort(unique(pdb_table[atom.chain == "A" & atom.type == "ATOM"]$atom.resno))
  num_residues <- length(residue_numbers)
  pdz_name_<-toupper(pdz_name)
  print(pdz_name_)
  
  # Create a temporary table for the secondary structure
  temporary_table <- data.table(
    WT_aa = pdbseq(struct), 
    Pos = residue_numbers, 
    Uniprot_pos_ref = residue_numbers+assays_info[pdz_name==pdz_name_ & assay=="abundance"]$start-1,
    secondary_structure = "loop"
  )
  
  # Annotate regions
  temporary_table[,secondary_structure:= "loop"]
  temporary_table[Pos %in% unlist(mapply(`:`, struct$helix$start, struct$helix$end)), 
                secondary_structure := "helix"]
  temporary_table[Pos %in% unlist(mapply(`:`, struct$sheet$start, struct$sheet$end)), 
                secondary_structure := "sheet"]
  
  # Add PDZ identifier
  temporary_table[, pdz := pdz_name]
  
  
  # read and add RSASA values [calculated in Jupyter notebook file]
  rsasa_filename<-paste0("data/annotations/RSASA_files/fold_", pdz_name,"_rsasa.csv")
  rsasa_file<-as.data.table(read.csv(rsasa_filename, sep="\t")[,-1])
 
  colnames(rsasa_file)[1]<-"Pos"
  len_prot<-nrow(rsasa_file)
  
  #STORE WT AA, PDZ and RSASA
  rsasa_temporary<-as.data.table(left_join(temporary_table, rsasa_file, by=c("Pos")))
  
  #get residue position in the structure (core/surface)
  rsasa_temporary[,structure_location:="surface"]
  rsasa_temporary[rsasa<0.25]$structure_location<-"core"
  
  # Append results
  secondary_structures <- rbind(secondary_structures, rsasa_temporary, fill = TRUE)

}
# save the results
write.csv(secondary_structures, paste0("data/annotations/tmp_files/secondary_structures_and_rsasa.csv"), row.names = FALSE)

```

## rsasa for files with the ligand
```{r}
# Load required files
secondary_structures<-as.data.table(read.csv(paste0("data/annotations/tmp_files/secondary_structures_and_rsasa.csv")))

# Initialize data.table to store secondary structures
rsasa_ligands_table <- data.table()
counter=1
for(lib in libraries_binding_names){
  print(paste0("reading: ", lib))
  
  rsasa_file<-as.data.table(read.csv(paste0("data/annotations/RSASA_files/fold_", lib, "_rsasa.csv"), sep="\t")[,-1])
  colnames(rsasa_file)[1]<-"Pos"
  len_prot<-nrow(rsasa_file)
  
  #STORE WT AA, PDZ and RSASA
  secondary_structures_tmp<-secondary_structures[pdz %in% str_extract(lib, pdz)]
  rsasa_temporary<-as.data.table(left_join(secondary_structures_tmp[,.(Pos, WT_aa)], rsasa_file, by=c("Pos")))
  colnames(rsasa_temporary)[3]<-"rsasa_w_ligand"
  rsasa_temporary[,library:=lib]
  
  # Append results
  rsasa_ligands_table <- rbind(rsasa_ligands_table, rsasa_temporary, fill = TRUE)
  
  counter=counter+1
}

# save the results
write.csv(rsasa_ligands_table, "data/annotations/tmp_files/rsasa_ligands_table.csv", row.names = FALSE)
```



# 3. LIGAND DISTANCES INFO

```{r}
# Initialize the distance matrix
distance_matrix <- data.table()

# Process each library
for (lib in libraries_binding_names) {
  print(paste0("reading: ", lib))
  ligand <- unlist(strsplit(assays_info[library==toupper(lib)]$ligand_sequence, ""))
  
  # Compute ligand distances
  pdb_dists_per_res <- doubledeepms__minimum_interchain_distances_from_PDB_perres(
    input_file = paste0("data/structures/alphafold3/", lib, ".pdb")
  )
  
  # Standardize table 
  pdb_dists_per_res[, Uniprot_pos_ref := Pos+assays_info[library==toupper(lib)]$start-1]
  pdb_dists_per_res[, library := lib]
  
  # Add binding interface information
  pdb_dists_per_res[, binding_interface_5A := scHAmin_ligand <= 5]
  
  # Append results to the distance matrix
  distance_matrix <- rbind(distance_matrix, pdb_dists_per_res)
}

# Save the complete distance matrix
write.csv(distance_matrix, file = paste0("data/annotations/tmp_files/ligand_distances.csv"), row.names = FALSE)
```


# 4. CONTACTS INFO
```{r}
# code from Taraneh's paper
contacts_complete_table<-data.table()
interactions_processed_complete_table<-data.table()
counter=1
for (lib in libraries_binding_names){
  pdz_seq<-assays_info[library==toupper(lib)]$pdz_sequence[1]
  sequence=unlist(strsplit(pdz_seq,""))
  
  ligand_seq<-assays_info[library==toupper(lib)]$ligand_sequence[1]
  lig_sequence=unlist(strsplit(ligand_seq,""))

  #load contacts results file
  print(paste0("reading: ", "data/annotations/contacts_files/", lib,"_contacts.tsv"))
  
  get_contacts_results <- as.data.table(read.delim(paste0("data/annotations/contacts_files/", lib,"_contacts.tsv"), skip=2, header=F))
  colnames(get_contacts_results)<-c("frame", "interaction_type", "atom_1", "atom_2")
  get_contacts_results[, library:=lib]

  #concatenate the raw table
  contacts_complete_table<-rbind(contacts_complete_table, get_contacts_results)
  
  #process the table in an easier format
  # Separate the last two columns by ":"
  get_contacts_results[, c("chain_1", "residue_name_1", "Pos_1", "atom_1_split") := tstrsplit(atom_1, ":", fixed = TRUE)]
  get_contacts_results[, c("chain_2", "residue_name_2", "Pos_2", "atom_2_split") := tstrsplit(atom_2, ":", fixed = TRUE)]
  
  # Convert positions to numeric for easier aggregation
  get_contacts_results[, Pos_1 := as.numeric(Pos_1)]
  get_contacts_results[, Pos_2 := as.numeric(Pos_2)]
  
  #invert the table and row bind (to have all interaction residues as residue_1)
  get_contacts_results_inverted<-get_contacts_results[,c("frame", "interaction_type", "atom_1", "atom_2")]
  get_contacts_results_inverted[, c("chain_1", "residue_name_1", "Pos_1", "atom_1_split") := tstrsplit(atom_2, ":", fixed = TRUE)]
  get_contacts_results_inverted[, c("chain_2", "residue_name_2", "Pos_2", "atom_2_split") := tstrsplit(atom_1, ":", fixed = TRUE)]
  
  get_contacts_results_complete<-rbind(get_contacts_results[,c("chain_1", "residue_name_1", "Pos_1", "atom_1_split","chain_2", "residue_name_2", "Pos_2", "atom_2_split", "interaction_type")], get_contacts_results_inverted[,c("chain_1", "residue_name_1", "Pos_1", "atom_1_split","chain_2", "residue_name_2", "Pos_2", "atom_2_split", "interaction_type")])
  
  get_contacts_results_complete[, ligand_backbone_contact:=atom_2_split %in% c("CA", "C", "N", "O")]
  get_contacts_results_complete[, pdz_backbone_contact:=atom_1_split %in% c("CA", "C", "N", "O")]
  
  # Aggregate by Pos_1 with a list of interaction_types
aggregated_interactions <- get_contacts_results_complete[
  , .(interaction_types = paste0(unique(interaction_type), collapse=","), interaction_types_all = paste0(interaction_type, collapse=","), 
      num_contacts = .N,
      pdz_backbone_contact = any(pdz_backbone_contact),
      pdz_sidechain_contact = any(pdz_backbone_contact==F),
      ligand_backbone_contact = any(ligand_backbone_contact),
      ligand_sidechain_contact = any(ligand_backbone_contact==F)), 
  by = .(chain_1, residue_name_1, Pos_1, chain_2, residue_name_2, Pos_2)]

  aggregated_interactions[, residue_1_letter := aa_mapping[residue_name_1]]
  aggregated_interactions[, residue_2_letter := aa_mapping[residue_name_2]]
  aggregated_interactions[,library:=lib]
  
  #concatenate the tables
  interactions_processed_complete_table<-rbind(interactions_processed_complete_table, aggregated_interactions)
  
counter=counter+1
}
interactions_processed_complete_table[,interaction_types:=paste0(interaction_types)]

#save a complete table
write.csv(interactions_processed_complete_table, paste0("data/annotations/tmp_files/contacts_table_complete.csv"))

```



# 6. ALIGNMENT
## [USED] T-coffee structural alignment of only my 6 PDZs
```{r warning=FALSE}
ids_my_pdzs=c("P78352_DLG4_HUMAN_PDZ-3", "P78352_DLG4_HUMAN_PDZ-2", "Q5T2W1_NHRF3_HUMAN_PDZ-2", "Q5T2W1_NHRF3_HUMAN_PDZ-1", "Q96RT1_ERBIN_HUMAN_PDZ-1")

alignment <- readAAStringSet(paste0("data/annotations/structural_alignment_tcoffee.fasta"), format = "fasta")

names(alignment)<-unlist(lapply(strsplit(names(alignment), " "), "[[", 1))

aln_table<-as.matrix(alignment)
aln_table_<-as.data.table(aln_table)
aln_table_$id<-unlist(lapply(strsplit(names(alignment), " "), "[[", 1))
rownames(aln_table_)<-rownames(aln_table)
aln_table<-as.data.table(aln_table)


# make a table with the alignment positions
PDZ_names<-c("PSD95_PDZ3",  "PSD95_PDZ2",  "SNTA1_PDZ1", "NHERF3_PDZ2", "NHERF3_PDZ1", "ERBIN_PDZ1")
aln_table$pdz_name<-PDZ_names
aln_table<-aln_table[pdz_name!="SNTA1_PDZ1"]
aln_table$pdz_name<-toupper(pdz_names)

# Convert to long format
long_alignment <- melt(aln_table, id.vars =c("pdz_name"), variable.name = "alignment_position", value.name = "WT")

# Add alignment positions as numeric
long_alignment[, structural_alignment_pos := extract_numeric(alignment_position)]

# Calculate raw_position (ignoring gaps '-')
long_alignment[, Pos := cumsum(WT != "-"), by = pdz_name]
long_alignment[WT == "-", Pos := NA]


#calculate conservation scores
conservation_scores<-conserv(as.matrix(aln_table), method="similarity", sub.matrix="blosum62")#[alignment_parts]
conservation_scores<-data.table(consv=conservation_scores, structural_alignment_pos=c(1:length(conservation_scores)))

#join alignment table and conservation table
long_alignment<-left_join(long_alignment, conservation_scores, by="structural_alignment_pos")

write.table(long_alignment, paste0("data/annotations/tmp_files/tcoffee_structural_alignment.csv"))

```


## sequence identity
```{r}
library(data.table)

# Convert data.table to matrix
aln_matrix <- as.matrix(aln_table)

# Number of sequences
num_seqs <- nrow(aln_matrix)

# Initialize identity matrix
identity_matrix <- matrix(NA, nrow = num_seqs, ncol = num_seqs, dimnames = list(1:num_seqs, 1:num_seqs))

# Compute pairwise percentage identities
for (i in 1:num_seqs) {
  for (j in i:num_seqs) {
    seq1 <- aln_matrix[i, ]
    seq2 <- aln_matrix[j, ]
    
    # Exclude gap positions ("-") from both sequences
    valid_positions <- (seq1 != "-") & (seq2 != "-")
    
    if (sum(valid_positions) > 0) {  # Avoid division by zero
      num_identical <- sum(seq1[valid_positions] == seq2[valid_positions])
      total_positions <- sum(valid_positions)
      percent_identity <- (num_identical / total_positions) * 100
    } else {
      percent_identity <- NA
    }
    
    identity_matrix[i, j] <- percent_identity
    identity_matrix[j, i] <- percent_identity  # Since the matrix is symmetric
  }
}

# Convert to data.table for better visualization
identity_dt <- as.data.table(identity_matrix, keep.rownames = "Seq1")
identity_dt

# Extract non-diagonal values (excluding self-comparisons)
non_diag_values <- identity_matrix[upper.tri(identity_matrix, diag = FALSE)]

# Compute the mean percentage identity
identity_matrix<-as.data.table(identity_matrix)
#write.table(identity_matrix, paste0("data/annotations/tmp_files/sequence_identity.csv"))

# Print result
print(mean(non_diag_values, na.rm = TRUE))
print(max(non_diag_values, na.rm = TRUE))
print(min(non_diag_values, na.rm = TRUE))
print(sd(non_diag_values, na.rm = TRUE))

```




## pairwise structural alignment RMSDs
this is pairwise using chimeraX (manually)
there are two measurements,
pruned RMSD: this is the RMSD computed only from the aligned residues
"all" RMSD: this is the RMSD computed from "all the residue pairs"

I am presenting the "all" RMSD here from the pairwise superimposition of the structures 
```{r}
list_rmsd_values<-c(1.956,2.925,1.854,1.948,2.433,1.948,2.155,5.654,2.151,3.082)
print(mean(list_rmsd_values))
print(median(list_rmsd_values))
print(max(list_rmsd_values))
print(min(list_rmsd_values))
```





# 5. BINDING INTERFACE (FROM CONTACTS)
```{r}
interactions_processed_complete_table<-as.data.table(read.csv("data/annotations/tmp_files/contacts_table_complete.csv"))[,-c("X")]
#in this case N is the number of ligand residues that it contacts
contacts_ligand_n<-interactions_processed_complete_table[chain_1=="A" & chain_2=="B", .N, by=.(library, Pos_1, residue_name_1)]
setnames(contacts_ligand_n, "Pos_1", "Pos")
contacts_ligand_n[,binding_interface_contacts:=T]
contacts_ligand_n[, pdz_name := toupper(sub("_[^_]+$", "", library))]

contacts_ligand_n<-left_join(contacts_ligand_n, long_alignment[,c( "Pos", "pdz_name", "structural_alignment_pos")], by=c("pdz_name", "Pos"))
core_bi_n<-contacts_ligand_n[, .N, by=.(structural_alignment_pos)][N>=7]
contacts_ligand_n[, core_bi_contacts:=structural_alignment_pos %in% core_bi_n$structural_alignment_pos]

write.csv(contacts_ligand_n, paste0("data/annotations/tmp_files/binding_interface_contacts.csv"))
```


# join all the metrics
```{r}
#binding data
distance_matrix[, pdz_name := toupper(sub("_[^_]+$", "", library))]
all_metrics<-merge(distance_matrix, long_alignment, by=c("pdz_name", "Pos"), all=T)
setnames(all_metrics, "WT", "WT_aa")

contacts_ligand_n[, pdz_name := toupper(sub("_[^_]+$", "", library))]
all_metrics<-as.data.table(merge(contacts_ligand_n[,-c("N", "residue_name_1")], all_metrics, by=c("library", "Pos", "pdz_name", "structural_alignment_pos"), all=T))

secondary_structures$pdz<-toupper(as.character(secondary_structures$pdz))
colnames(secondary_structures)[5]<-"pdz_name"
all_metrics<-as.data.table(merge(all_metrics, secondary_structures, by=c("WT_aa", "pdz_name", "Pos"), all=T))

#add abundance
all_metrics_ab<-as.data.table(merge(long_alignment, secondary_structures, by=c("pdz_name", "Pos"), all=T))
all_metrics_ab[,library:=tolower(pdz_name)]

#join all together
all_metrics<-rbind(all_metrics, all_metrics_ab, fill=T)

write.table(all_metrics, paste0("data/annotations/structure_metrics.csv"))
```
