---
title: "Normalizing fitness"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---


# initializing
This script will 
1. normalize the fitness values between blocks and libraries.
2. Plots of the overlapping regions (Figure S1B)
3. Choose alternative backgrounds as WT when there are no singles in some of the blocks.

```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

# load some variables
load('source_variables.RData')
```

# libraries used for the analysis
```{r, warning=F, results='hide', message=FALSE}
library(car)
library(systemfonts)
library(plyr)
library(ggiraph)
library(ggiraphExtra)
library(rstatix)
library(data.table)
library(dplyr)
library(seqinr)
library(stringr)
library(ggplot2)
library(Matrix)
library(viridis)
library(ggpubr)
library(readxl)
library(stringi)
library(GGally)
library("readxl")
library(modelr)
library(cowplot)
library(Hmisc)
library(ggrepel)
library(dplyr) 
library(ddpcr)
library(ggpointdensity)
library(ggpmisc)
```


# normalize between blocks
I am showing the entire distributions of stop and synonymous variants
Then I am taking the stops of the overlapping regions + overall synonymous and normalyzing by this

```{r}
library(ggridges)
syn_stops_plots<-list()
syn_stops_plots_norm<-list()
all_variants_all<-as.data.table(read.table(paste0("tmp_tables/fitness_all_variants.txt"),sep=",", header = T))
all_variants_all[is.na(WT), WT:=F]
all_variants_all[,category:="missense"]
all_variants_all[Nham_aa==0 & WT==F, category:="synonymous"]
all_variants_all[STOP==T, category:="stop"]

c=1
for(lib in libraries_names){
  print(lib)
  if(is.na(str_extract(lib, "psd95_pdz3"))){
    both_blocks<-all_variants_all[library==lib]
    
    all_syn<-both_blocks[category=="synonymous"]
    all_stops<-both_blocks[category=="stop"]
    
    length_aa<-nchar(all_syn$aa_seq[1]) #total length of the PDZ
    
    #get the position of each first stop codon
    stop_positions<-unlist(lapply(sapply("\\*", gregexpr, lapply(all_stops$aa_seq, substr, 1, length_aa-1)),"[[",1))
    
    #add this to the stops table
    all_stops[,stop_pos:=stop_positions[!stop_positions %in% length_aa]] 
    b1_s<-all_stops[block==1]
    b2_s<-all_stops[block==2]
    
    # get the overlapping stops present in both blocks
    overlap_s<-all_stops[stop_pos %in% b1_s$stop_pos & stop_pos %in% b2_s$stop_pos,]
    syn_stops_overlap<-rbind(overlap_s, all_syn,fill=TRUE)
    
  
    p<-ggplot(syn_stops_overlap[category!="missense"])+
      geom_density_ridges(aes(x=fitness, y=as.factor(block), fill=category),quantile_lines = TRUE,quantiles = 2, linewidth=1, alpha=0.5)+
      ylab("block")+
      theme_classic()+
      ggtitle(paste0(libraries_names[which(libraries==lib)], "\noverlap variants"))+
      scale_fill_manual(values=c("red", "green2"))+theme(legend.position = "none")
    
    syn_stops_plots[[c]]<-p
    
    # prepare data blocks to normalize
    median_stops_b1<-median(syn_stops_overlap[category=="stop" & block==1]$fitness)
    median_stops_b2<-median(syn_stops_overlap[category=="stop" & block==2]$fitness)
    
    if(length(unique(syn_stops_overlap[block==1]$category))==1){median_syn_b1<-0}
    else{median_syn_b1<-median(syn_stops_overlap[category=="synonymous" & block==1]$fitness)}
    if(length(unique(syn_stops_overlap[block==2]$category))==1){median_syn_b2<-0}
    else{median_syn_b2<-median(syn_stops_overlap[category=="synonymous" & block==2]$fitness)}
    data<-data.table(b1=c(median_stops_b1, median_syn_b1), b2=c(median_stops_b2, median_syn_b2))
    
    #build model
    model <- lm(b1 ~ b2, data = data)
    
    #modify block2 to block1
    b2_norm<-as.vector(predict(model, data.frame(b2=both_blocks[block==2]$fitness)))
    b2_std_norm<-as.vector(both_blocks[block==2]$sigma*model$coefficients[2])
    
    #save the data in the fitness_norm_blocks column
    both_blocks$fitness_norm_blocks<-both_blocks$fitness
    both_blocks[block==2]$fitness_norm_blocks<-b2_norm
    both_blocks$std_norm_blocks<-both_blocks$sigma
    both_blocks[block==2]$std_norm_blocks<-b2_std_norm
      
    #save the normalized fitness
    write.table(both_blocks, paste0("tmp_tables/normalize_fitness/blocks_normalized_fitness_", lib, ".csv"))
    
    p<-ggplot(both_blocks[category!="missense"])+
      geom_density_ridges(aes(x=fitness_norm_blocks, y=as.factor(block), fill=category),quantile_lines = TRUE,quantiles = 2, linewidth=1, alpha=0.5)+
      ylab("block")+
      theme_classic()+
      ggtitle(paste0(libraries_names[which(libraries==lib)], "\nall variants"))+
      scale_fill_manual(values=c("red", "green2", "blue"))+theme(legend.position = "none")
    
    syn_stops_plots_norm[[c]]<-p
    c=c+1
  }
  else{
    #just save the table (only one block = already normalized)
    both_blocks<-all_variants_all[library==lib]
    
    all_syn<-both_blocks[category=="synonymous"]
    all_stops<-both_blocks[category=="stop"]
    both_blocks$fitness_norm_blocks<-both_blocks$fitness
    write.table(both_blocks, paste0("tmp_tables/normalize_fitness/blocks_normalized_fitness_", lib, ".csv"))
  }
}

plot_grid(plotlist=syn_stops_plots, ncol=5)
plot_grid(plotlist=syn_stops_plots_norm, ncol=5)
```

# normalize between libraries
I will normalize all to 762_809 as it had a very nice dynamic range and fitness correlations

```{r}
#reference data
syn_stops_plots_norm<-list()
reference_lib<-"psd95_pdz2_nmdar2a"
block_norm_fitness_dt_ref<-as.data.table(read.table(paste0("tmp_tables/normalize_fitness/blocks_normalized_fitness_", reference_lib, ".csv")))
ref_stops_median<-median(block_norm_fitness_dt_ref[category=="stop"]$fitness_norm_blocks, na.rm = T)
ref_syn_median<-median(block_norm_fitness_dt_ref[category=="synonymous"]$fitness_norm_blocks, na.rm = T)
tables_all<-data.table()

plots_norm<-list()
c=1
for(lib in libraries_names){
  pdz_=substr(lib, 1,3)
  if(lib!=reference_lib){
    print(lib)
    
    #prepare data
    block_norm_fitness_dt<-as.data.table(read.table(paste0("tmp_tables/normalize_fitness/blocks_normalized_fitness_", lib, ".csv")))
    stops_median<-median(block_norm_fitness_dt[category=="stop"]$fitness_norm_blocks)
    syn_median<-median(block_norm_fitness_dt[category=="synonymous"]$fitness_norm_blocks)
    if(is.na(syn_median)){syn_median<-0}
    data<-data.table(f1=c(ref_stops_median, ref_syn_median), f2=c(stops_median, syn_median))
    
    #prepare model
    model <- lm(f1 ~ f2, data = data)
    
    #modify library2 to library1
    f2_norm<-as.vector(predict(model, data.frame(f2=block_norm_fitness_dt$fitness_norm_blocks)))
    f2_std_norm<-as.vector(block_norm_fitness_dt$std_norm_blocks*model$coefficients[2])
    
    block_norm_fitness_dt$fitness_norm<-f2_norm
    block_norm_fitness_dt$std_norm<-f2_std_norm
    block_norm_fitness_dt$library<-lib
    
    #save new data
    write.table(block_norm_fitness_dt, paste0("tmp_tables/normalize_fitness/libraries_normalized_fitness_", lib, ".csv"))
    
    c=c+1
    
    tables_all<-rbind(tables_all, block_norm_fitness_dt, fill=T)
    
  }
  else{ #reference lib: just save the table without normalizing
    block_norm_fitness_dt<-as.data.table(read.table(paste0("tmp_tables/normalize_fitness/blocks_normalized_fitness_", lib, ".csv")))
    block_norm_fitness_dt$fitness_norm<-block_norm_fitness_dt$fitness_norm_blocks
    block_norm_fitness_dt$std_norm<-block_norm_fitness_dt$std_norm_blocks
    block_norm_fitness_dt$library<-lib
    
    write.table(block_norm_fitness_dt, paste0("tmp_tables/normalize_fitness/libraries_normalized_fitness_", lib, ".csv"))
    tables_all<-rbind(tables_all, block_norm_fitness_dt, fill=T)
  }
}

write.table(tables_all, paste0("tmp_tables/libraries_normalized_fitness_all.csv"))
```

# plot normalized libraries
```{r}
all_normalized<-data.table(read.table(paste0("tmp_tables/libraries_normalized_fitness_all.csv")))
```

```{r fig.width=5, fig.height=7, warning=F}
library(ggridges)

#change library names from codes to actual names:
all_normalized$library<-lib_code_to_name(all_normalized$library, libraries_names, libraries_names_plots)

p<-ggplot(all_normalized)+
      geom_density_ridges(aes(x=fitness_norm, y=library, fill=category),quantile_lines = TRUE,quantiles = 2, linewidth=1, alpha=0.4)+
      theme_classic(base_size=16)+
      xlim(c(-3,1))+
      #ggtitle("All libraries normalized")+
  theme(axis.title.y = element_blank(),
        legend.position = "top",
        legend.title = element_blank())+
  xlab("fitness normalized")+
      scale_fill_manual(values=c("blue","red", "green3"))#+theme(legend.position = "none")
p

p2<-ggplot(all_normalized[all_normalized$library %in% c("ERBIN_PDZ1 abundance", "ERBIN_PDZ1 binding CRIPT"),])+
      geom_density_ridges(aes(x=fitness_norm, y=factor(library, levels=c("ERBIN_PDZ1 abundance", "ERBIN_PDZ1 binding CRIPT"), labels=c("ERBIN PDZ1\nabundance", "ERBIN PDZ1\nbinding CRIPT")), fill=category),quantile_lines = TRUE,quantiles = 2, linewidth=1, alpha=0.4)+
      theme_classic(base_size=16)+
      xlim(c(-3,1))+
      #ggtitle("All libraries normalized")+
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        axis.title.x = element_text(size=13),
        legend.position = "top",
        legend.title = element_blank(),
        legend.key.size = unit(0.5, 'cm'),
        legend.text = element_text(size=10),
        legend.justification = "right")+
  xlab("fitness normalized")+
      scale_fill_manual(values=c("blue","red", "green3"))+#+theme(legend.position = "none")
  guides(fill=guide_legend(nrow=1,byrow=TRUE))
p2
ggsave(paste0("Figs/Fig1/Fig1I_distributions_examples.png"), width=3.2, height=3.7 )



```



# libraries that do not have singles in one block
the main problem here is that some blocks do not have singles because the background didn't codify for the wt. 
I create a table with the libraries that have a problem and the new background that will substitute the singles.

```{r}
#check which are the libraries that do not have singles
lib_changed<-c()
block_changed<-c()
single_positions<-c()
new_background_wt<-c()
fitness_<-c()
fitness_norm_<-c()
for(lib in libraries_names){
  print(lib)
  norm_fitness_dt<-as.data.table(read.table(paste0("tmp_tables/normalize_fitness/libraries_normalized_fitness_", lib, ".csv")))[Nham_aa==1]
  expected<-nchar(norm_fitness_dt$aa_seq[1])*19
  if(nrow(norm_fitness_dt)<expected/2){
    print("changing lib")
    lib_changed<-c(lib_changed, lib)
    for(b in 1:2){
      if(nrow(norm_fitness_dt[block==b])<30){
        print(paste0("changing block", b))
        block_changed<-c(block_changed,b)
        norm_fitness_dt_<-norm_fitness_dt[block==b]
        norm_fitness_dt_[,Pos:=as.numeric(gsub(".*?([0-9]+).*", "\\1", mut_id))]
        single_positions<-c(single_positions,nrow(norm_fitness_dt_))
        mildest_backgr<-norm_fitness_dt_[abs(fitness_norm)==min(abs(norm_fitness_dt_$fitness_norm))]
        new_background_wt<-c(new_background_wt, mildest_backgr$mut_id)
        fitness_<-c(fitness_, mildest_backgr$fitness)
        fitness_norm_<-c(fitness_norm_, mildest_backgr$fitness_norm)
      }
    }
  }
}

fitness_tables_modifications<-data.table("library"=lib_changed, "block"=block_changed, "num_singles"=single_positions,"new_wt"=new_background_wt, "fitness"=fitness_, "fitness_norm"=fitness_norm_)
print(fitness_tables_modifications)
write.csv(fitness_tables_modifications, paste0("tmp_tables/fitness_tables_modifications.csv"))

```


# choosing a new WT
For the libraries that do not have singles in one of the blocks, I will take as new wt the mildest background. Closest to 0.
```{r}
all_normalized<-data.table(read.table(paste0("tmp_tables/libraries_normalized_fitness_all.csv")))
```

```{r}
library(tidyr)
modifications_libraries<-as.data.table(read.table(paste0("tmp_tables/fitness_tables_modifications.csv"), sep=",", header=T))
all_fitness_modified<-data.table()

for(lib in libraries_names){
  if(lib %in% modifications_libraries$library){
    #initialize variables
    new_wt<-modifications_libraries[library==lib]$new_wt
    block_to_modify<-modifications_libraries[library==lib]$block
    print(paste0("problem with library ", lib))
    print(paste0("problem with block ", block_to_modify))
    
    #load data
    unmodified_block<-as.data.table(all_normalized)[library==lib & !block==block_to_modify & Nham_aa<2]
    block_to_modify_dt<-as.data.table(all_normalized)[library==lib & block==block_to_modify]
    block_to_modify_dt[,Pos:=as.numeric(gsub(".*?([0-9]+).*", "\\1", mut_id))]
    mildest_backgr<-new_wt
    
    # subset to the doubles with the mildest background
    block_to_modify_dt$mut_id_original<-block_to_modify_dt$mut_id
    block_to_modify_dt<-as.data.table(block_to_modify_dt[Nham_aa==2] %>% separate(mut_id_original, c("mut1", "mut2"), ","))
    new_background_dt<-block_to_modify_dt[mut1==mildest_backgr | mut2==mildest_backgr]
    new_background_wt<-as.data.table(all_normalized)[library==lib & block==block_to_modify & extract_numeric(mut_id)==extract_numeric(mildest_backgr)] 
    new_background_wt[mut_id==mildest_backgr, WT:=T]
    new_background_wt[,modified_background:=T]
    new_background_wt[,new_background:=mildest_backgr]
  
    #modify the table and make these the new singles
    new_background_dt[,modified_background:=T]
    new_background_dt[,new_background:=mildest_backgr]
    new_background_dt[, old_mut_id := mut_id]
    new_background_dt[, mut_id := ifelse(mut1 != mildest_backgr, mut1, mut2)]
    
    #now join the two sets
    new_background_dt_complete<-rbind(new_background_dt, new_background_wt, fill=TRUE)
    both_blocks_modified_background<-rbind(unmodified_block, new_background_dt_complete, fill=TRUE)
    
    #save the original vs modified table
    write.csv(both_blocks_modified_background, paste0("tmp_tables/singles_fitness/singles_modified_background_", lib, ".csv"))
    
    all_fitness_modified<-rbind(all_fitness_modified, both_blocks_modified_background)
    
    write.csv(as.data.table(all_normalized)[library==lib & Nham_aa<2], paste0("tmp_tables/singles_fitness/singles_original_background_", lib, ".csv"))
    
    #print number of singles originally vs the modified one
    print(paste0("original number of singles:", nrow(as.data.table(all_normalized)[library==lib & Nham_aa<2])))
    print(paste0("modified number of singles:", nrow(both_blocks_modified_background)))
    print(paste0("Expected number of singles:",nchar(both_blocks_modified_background$aa_seq[1])*20, "+overlap"))
  }
  else{
    print(paste0("NO problem with library ", lib))
    print(paste0("original number of singles:", nrow(as.data.table(all_normalized)[library==lib & Nham_aa<2])))
    write.csv(as.data.table(all_normalized)[library==lib & Nham_aa<2], paste0("tmp_tables/singles_fitness/singles_original_background_", lib, ".csv"))
  }
}


```


# plot overlapping regions after modification
```{r, eval=F}
all_normalized<-data.table(read.table(paste0("tmp_tables/libraries_normalized_fitness_all.csv")))
```

```{r}
modifications_libraries<-as.data.table(read.table(paste0("tmp_tables/fitness_tables_modifications.csv"), sep=",", header=T))

modified_overlapping_plots<-list()
all_fitness_modified<-data.table()
c=1

for(lib in libraries_names){
  col="black"
  if(is.na(str_extract(lib, "psd95_pdz3"))){
  
  if (lib %in% modifications_libraries$library){modified_table<-as.data.table(read.table(paste0("tmp_tables/singles_fitness/singles_modified_background_", lib, ".csv"), sep=",", header=T))
  col="tomato"}
  else{modified_table<-as.data.table(all_normalized)[library==lib]}
    
  modified_table[,mutation:=substr(mut_id, nchar(mut_id), nchar(mut_id))]
  ggplot(modified_table)+
    geom_tile(aes(x=Pos, y=mutation, fill=fitness_norm))
  
  #join all final used data (all modified)
  all_fitness_modified<-rbind(all_fitness_modified, modified_table, fill=TRUE)
  
  # identify overlapping mutations
  overlapping_mutations<-as.list(modified_table[duplicated(modified_table$mut_id) & modified_table$mut_id!="", "mut_id"])$mut_id
  
  # extract overlapping regions
  overlapping_blocks_tmp<-modified_table[modified_table$mut_id %in% overlapping_mutations ,c("aa_seq", "Nham_aa", "block", "mut_id", "fitness_norm", "sigma")]
  overlapping_b1<-overlapping_blocks_tmp[block==1,]
  overlapping_b2<-overlapping_blocks_tmp[block==2,]
  overlapping_blocks<-merge(overlapping_b1, overlapping_b2, by=c("mut_id"), suffixes=c(".b1", ".b2"))
  
  # Generate correlation plot
  p_corr_overlap<-ggplot(overlapping_blocks,aes(x=fitness_norm.b1, y=fitness_norm.b2))+
        theme_classic()+
        xlab("fitness block1") + 
        ylab("fitness block2") + 
        geom_hline(yintercept=0, color="grey")+
        geom_vline(xintercept=0, color="grey") +
        geom_abline(color="grey")+
        geom_point(alpha=0.5, size=1, color=col)+ 
        coord_fixed(ratio = 1, xlim = c(-1.7, 0.5), ylim = c(-1.7, 0.5))+
        stat_poly_eq(size=3) +
        ggtitle(sub(" ", "\n", libraries_names[which(libraries==lib)]))+ 
        theme(plot.title = element_text(size=6), axis.text.x = element_text(angle=90, size=7,vjust = 0.5, hjust=1), axis.text.y = element_text(size=7), axis.title.x = element_text(size=7), axis.title.y = element_text(size=7))

  modified_overlapping_plots[[c]]<-p_corr_overlap
  c=c+1
  all_fitness_modified<-rbind(all_fitness_modified, modified_table, fill=TRUE)
  
  }
  else{modified_table<-as.data.table(all_normalized)[library==lib]
   all_fitness_modified<-rbind(all_fitness_modified, modified_table, fill=TRUE)}
}
```

```{r}
all_fitness_modified[is.na(modified_background), modified_background:=F]


#for non modified backgrounds, it has to be comparison of singles
all_fitness_modified_singles_non_modified<-all_fitness_modified[modified_background==F & Nham_aa==1]
all_fitness_modified_singles_non_modified[, Pos:=extract_numeric(mut_id)]
all_fitness_modified_singles_non_modified[, mutation:=substr(mut_id, nchar(mut_id), nchar(mut_id))]

#for modified backgrounds, it has to be comparison of NEW singles
all_fitness_modified_singles_modified<-all_fitness_modified[modified_background==T & Nham_aa>=1]
all_fitness_modified_singles_modified[, Pos:=extract_numeric(mut_id)]
all_fitness_modified_singles_modified[, mutation:=substr(mut_id, nchar(mut_id), nchar(mut_id))]

# now I join both cases
all_fitness_modified_singles<-rbind(all_fitness_modified_singles_modified, all_fitness_modified_singles_non_modified)

all_fitness_modified_wide<-reshape(all_fitness_modified_singles[,c("library", "block", "fitness_norm", "std_norm", "mutation", "Pos")], idvar = c("library", "mutation", "Pos"), timevar = "block", direction = "wide")
# now I add the info for the modified backgrounds
all_fitness_modified_wide[, modified_background:=library %in% modifications_libraries$library]


all_fitness_modified_wide$library_name<-gsub("_", "-", gsub(" binding ", "\n", (unlist(lapply(all_fitness_modified_wide$library, lib_code_to_name, libraries_names, libraries_names_plots)))))
all_fitness_modified_wide$library_name<-gsub(" abundance", "\nabundance", all_fitness_modified_wide$library_name)

ggplot(all_fitness_modified_wide[!library %in% c("psd95_pdz3", "psd95_pdz3_cript")], aes(x=fitness_norm.1, y=fitness_norm.2, color=modified_background))+
  scale_color_manual(values=c("black", "red"))+
  stat_cor(aes(label = paste(..rr.label.., sep = "~`,`~")), label.y = 1, label.x=-1.9, color="black", size=4)+
  geom_point(alpha=0.5)+
  geom_errorbar(aes(ymin = fitness_norm.2 - std_norm.2, ymax = fitness_norm.2 + std_norm.2), width = 0.1, alpha = 0.3) +
  geom_errorbar(aes(xmin = fitness_norm.1 - std_norm.1, xmax = fitness_norm.1 + std_norm.1), width = 0.1, alpha = 0.3) +
  geom_abline()+
  facet_wrap(~library_name, nrow=5)+
  theme_classic()+
  xlab("normalized fitness block1")+
  ylab("normalized fitness block2")+
  theme(legend.position="none")+
  ylim(c(-2.2, 1.3))+
  theme(legend.position="none",
        strip.text.x = element_text(size = 8, face = "bold"),
        strip.background = element_rect(color="grey", fill="white", size=1.5, linetype="solid"
     ))

ggsave(paste0("Figs/Fig1/FigS1B_normalized_overlapping.png"), width=3, height=8)

```


# save a final table with all the normalizations for supplementary material
load the working table
```{r}
all_normalized<-data.table(read.table(paste0("tmp_tables/libraries_normalized_fitness_all.csv")))
```

clean and save the table
```{r}
all_normalized<-as.data.table(all_normalized)[, -c("V1", "STOP_readthrough", "indel", "Nham_nt", "nt_seq")]
setnames(all_normalized, "fitness1_uncorr", "fitness_r1")
setnames(all_normalized, "fitness2_uncorr", "fitness_r2")
setnames(all_normalized, "fitness3_uncorr", "fitness_r3")
setnames(all_normalized, "fitness_norm_blocks", "fitness_norm_between_blocks")
all_normalized$library<-lib_code_to_name(all_normalized$library, libraries, libraries_names)

all_normalized<-all_normalized %>%
  relocate(block)

all_normalized<-all_normalized %>%
  relocate(library)
fwrite(all_normalized, "supplementary_tables/Supplementary_table4.txt")
```



