---
title: "all_ddg tables and ddG heatmaps"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---

# initializing 
```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

# load some variables
load('source_variables.RData')

library(data.table)
aln_table<-as.data.table(read.table(paste0("/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis/01_structures/3DstructureMetrics/01-alignment_positions/", "tcoffee_structural_alignment.csv")))

```

## libraries used for the analysis
```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(bio3d)
library(dplyr)
library(stringr)
library(readxl)
library(tidyr)
library(cowplot)
library(ggpmisc)
library(ggpubr)
library(scales)
library(ggpointdensity)

sessionInfo()

```


# create simpler ∆∆G tables
```{r}
all_weights_data <- as.data.table(fread(paste0("tmp_tables/weights_mochi_all.csv"), sep = "\t"))[,-c("V1")]
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))

all_weights_data[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]
all_weights_data[,WT_aa:=substr(id,1,1)]
all_weights_data<-all_weights_data[id!="WT",] #I do not want the WTs

all_weights_data[assay=="folding",library:=sub("^([^_]+_[^_]+).*", "\\1", library)]

#aggregate to folding
pdb_metrics_dt[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]
all_weights_data_f_complete<-left_join(all_weights_data[assay=="folding"], pdb_metrics_dt[nchar(pdz_name)==nchar(library),c("secondary_structure", "rsasa", "structure_location", "consv","pdz", "Pos", "WT_aa", "structural_alignment_pos")], by=c("pdz", "Pos", "WT_aa"))

#aggregate to binidng
all_weights_data_b_complete<-left_join(all_weights_data[assay=="binding"], pdb_metrics_dt[nchar(pdz_name)!=nchar(library)], by=c("pdz", "Pos", "WT_aa", "library"))

#join both
all_weights_data_complete<-rbind(all_weights_data_b_complete, all_weights_data_f_complete,fill=TRUE)


setnames(all_weights_data_complete, "mean_kcal.mol", "ddg")
setnames(all_weights_data_complete, "std_kcal.mol", "std_ddg")
setnames(all_weights_data_complete, "ci95_kcal.mol", "ci95_ddg")
write.csv(all_weights_data_complete, paste0("tmp_tables/all_ddg_table.csv"))
```

# saving a clean version of this for supplementary table 5
```{r}
# saving a clean version of this for supplementary table 5
supplementary_table_5<-all_weights_data_complete[, -c("pdz", "core_bi_contacts", "HAmin_ligand", "X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "Pos_ref.x", "Pos_ref.y", "closest_ligand_pos", "closest_ligand_aa", "pdz_name", "alignment_position", "WT", "Pos_ref", "trait_name")]
supplementary_table_5[is.na(binding_interface_contacts),binding_interface_contacts:=F]
supplementary_table_5[, Mut:=substr(id, nchar(id), nchar(id))]

start_positions<-c(311, 311, 157, 157, 157, 123, 123, 1,1,1,1321,1321)
supplementary_table_5[, Pos_ref:=0]
c=1
for (lib in libraries){
  supplementary_table_5[library==lib, Pos_ref:=Pos+start_positions[c]-1]
  c=c+1
}

supplementary_table_5$library<-lib_code_to_name(supplementary_table_5$library, libraries, libraries_names)

supplementary_table_5<-supplementary_table_5[,.(library, assay, id, WT_aa, Pos, Pos_ref, Mut, ddg, std_ddg, ci95_ddg, structural_alignment_pos, binding_interface_contacts, binding_interface_5A, scHAmin_ligand, consv, secondary_structure, rsasa, structure_location)]
setnames(supplementary_table_5, "consv", "sequence_conservation")
fwrite(supplementary_table_5, "supplementary_tables/Supplementary_table5.txt")
```





# create medians ddg table
```{r}
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))
pdb_metrics_dt[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]

aggregated_data <- all_weights_data_complete[
  , .(median_ddg = median(ddg, na.rm = TRUE)),  # Calculate median of ddg
  by = .(library, structural_alignment_pos, Pos,assay, WT_aa, pdz)
]

all_weights_data_complete_median<-left_join(aggregated_data, pdb_metrics_dt, by=c("pdz", "Pos", "structural_alignment_pos", "WT_aa", "library"))

write.csv(all_weights_data_complete_median, paste0("tmp_tables/all_median_ddg_table.csv"))
```

# complete supplementary5 with the medians
Now I add the medians to the final supplementary table
```{r}
supplementary_table_5<-left_join(supplementary_table_5, all_weights_data_complete_median[,.(library, Pos, median_ddg)], by=c("library", "Pos"))

fwrite(supplementary_table_5, "supplementary_tables/Supplementary_table5.txt")
```


# check that all is good by plotting the heatmaps w BI highlighted
```{r, fig.width=20, fig.height=4}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))
gg_medians_alignment_heatmap(all_ddg_table)+geom_tile(data = all_ddg_table[binding_interface_contacts==T], fill = "transparent", color="green", size = 0.7)+theme(axis.text.x=element_text(angle=90))

gg_medians_alignment_heatmap(all_ddg_table[nchar(library)==nchar(pdz)], assay_plotted = "folding")+theme(axis.text.x=element_text(angle=90))
```

# ddG heatmaps
```{r}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_ddg_table.csv")))
all_ddg_table[,Mut:=substr(id, nchar(id), nchar(id))]
all_ddg_table$class<-all_ddg_table$structure_location
all_ddg_table[binding_interface_5A==T]$class<-"BI"
all_ddg_table[assay=="binding" & structural_alignment_pos %in% c(24,25,26)]$class<-"CBL"

pdb_metrics_dt$pdz<-as.character(pdb_metrics_dt$pdz)
all_ddg_table$pdz<-as.character(all_ddg_table$pdz)


for(lib in libraries_names){
  all_ddg_table_subset<-all_ddg_table[library==lib]
  pdz<-sub("^([^_]+_[^_]+).*", "\\1", lib)
  
  start_positions<-c(311, 311, 157, 157, 157, 123, 123, 1,1,1,1321,1321)
  all_ddg_table_subset[, Pos_ref:=0]
  all_ddg_table_subset$Pos_ref<-as.numeric(all_ddg_table_subset$Pos_ref)
  all_ddg_table_subset[, Pos_ref:=Pos+start_positions[which(libraries_names==lib)]-1]
  
  
  all_ddg_table_subset[,Pos:=Pos_ref]
  if(nchar(pdz)!=nchar(lib)){p<-gg_ddg_heatmap(all_ddg_table_subset[!is.na(Pos)], aa_order2, colors_aa, gsub("_", "-", gsub(" binding ", " | ", libraries_names[which(libraries==lib)])))}
  
  else{
    p<-gg_ddg_heatmap(all_ddg_table_subset[!is.na(Pos)], aa_order2, colors_aa, gsub("_", "-", gsub(" binding ", " | ", libraries_names_plots[which(libraries_names==lib)])))+
  theme(legend.position = "right")
  }
    
  p+theme(legend.position="none", axis.text.x=element_text(size=22, face="bold"),axis.text.y=element_text(size=20, face="bold"))
  ggsave(paste0("Figs/Fig2/ddg_heatmaps/ddG_", lib, "_heatmap.png"), width=18, height=6)
}
```

# structure shapes
create the heatmap
```{r}
all_median_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))
#merge(all_median_ddg_table, aln_table)

for(lib in libraries_names){
  pdz<-sub("^([^_]+_[^_]+).*", "\\1", lib)
  
  if(pdz %in% c("psd95_pdz2", "erbin_pdz1")){point_height=1.05}
  if(pdz %in% c("psd95_pdz3")){point_height=1.03}
  if(pdz %in% c("nherf3_pdz2")){point_height=1.11}
  if(pdz %in% c("nherf3_pdz1")){point_height=1.09}
  
  ddg_subset<-all_median_ddg_table[library==lib]
  
  p<-ggplot(ddg_subset,aes(x=as.factor(Pos), y=factor(library)))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=median_ddg), color="black")+ 
    theme_void(base_size = 16)+
    scale_x_discrete(limits = seq(1, max(ddg_subset$Pos, na.rm = T), 1))+
    scale_fill_gradient2(limits=c(-0.8932919, 3.670025),"median ∆∆G",high="red",mid="grey90",low="blue",midpoint=0, na.value = "white")+
    theme( plot.title = element_text(size=20))+
    #guides(fill=guide_legend(title="weighted mean \nfitness normalized"))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           legend.position = "none")+#,
           #strip.text = element_text(size=10)) + 
    ggtitle(paste0(""))+
    geom_point(data=ddg_subset[binding_interface_contacts==T], aes(y=point_height), shape=21, fill="red", size=2) 
  
  
  ggsave(paste0("Figs/Fig2/colored_heatmap/", lib, ".png"), width=10, height=1.5)
}
```
Create the shaped heatmap
```{r}
library(magick)
# do not execute for now
for(lib in libraries_names){
  print(lib)
  path_image<-paste0("Figs/Fig2/colored_heatmap/", lib,".png")
  path_shape<-paste0("data/shapes/", lib, "_shape.png")
  out_path<-paste0("Figs/Fig2/colored_shapes/", lib, "_median_ddg.png")
  crop_heatmap_structure_shape_outline(path_image, path_shape, out_path)
}
```


