---
title: "all_ddg tables and ddG heatmaps"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---

# initializing 
```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

# load some variables
load('source_variables.RData')

library(data.table)
aln_table<-as.data.table(read.table(paste0("/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis/01_structures/3DstructureMetrics/01-alignment_positions/", "tcoffee_structural_alignment.csv")))

```

## libraries used for the analysis
```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(bio3d)
library(dplyr)
library(stringr)
library(readxl)
library(tidyr)
library(cowplot)
library(ggpmisc)
library(ggpubr)
library(scales)
library(ggpointdensity)

sessionInfo()

```


# create simpler ∆∆G tables
```{r}
all_weights_data <- as.data.table(fread(paste0("tmp_tables/weights_mochi_all.csv"), sep = "\t"))[,-c("V1")]
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))

all_weights_data[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]
all_weights_data[,WT_aa:=substr(id,1,1)]
all_weights_data<-all_weights_data[id!="WT",] #I do not want the WTs

all_weights_data[assay=="folding",library:=sub("^([^_]+_[^_]+).*", "\\1", library)]

#aggregate to folding
pdb_metrics_dt[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]
all_weights_data_f_complete<-left_join(all_weights_data[assay=="folding"], pdb_metrics_dt[nchar(pdz_name)==nchar(library),c("secondary_structure", "rsasa", "structure_location", "consv","pdz", "Pos", "WT_aa", "structural_alignment_pos")], by=c("pdz", "Pos", "WT_aa"))

#aggregate to binidng
all_weights_data_b_complete<-left_join(all_weights_data[assay=="binding"], pdb_metrics_dt[nchar(pdz_name)!=nchar(library)], by=c("pdz", "Pos", "WT_aa", "library"))

#join both
all_weights_data_complete<-rbind(all_weights_data_b_complete, all_weights_data_f_complete,fill=TRUE)


setnames(all_weights_data_complete, "mean_kcal.mol", "ddg")
setnames(all_weights_data_complete, "std_kcal.mol", "std_ddg")
setnames(all_weights_data_complete, "ci95_kcal.mol", "ci95_ddg")
write.csv(all_weights_data_complete, paste0("tmp_tables/all_ddg_table.csv"))
```

# saving a clean version of this for supplementary table 5
```{r}
# saving a clean version of this for supplementary table 5
supplementary_table_5<-all_weights_data_complete[, -c("pdz", "core_bi_contacts", "HAmin_ligand", "X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "Pos_ref.x", "Pos_ref.y", "closest_ligand_pos", "closest_ligand_aa", "pdz_name", "alignment_position", "WT", "Pos_ref", "trait_name")]
supplementary_table_5[is.na(binding_interface_contacts),binding_interface_contacts:=F]
supplementary_table_5[, Mut:=substr(id, nchar(id), nchar(id))]

start_positions<-c(311, 311, 157, 157, 157, 123, 123, 1,1,1,1321,1321)
supplementary_table_5[, Pos_ref:=0]
c=1
for (lib in libraries){
  supplementary_table_5[library==lib, Pos_ref:=Pos+start_positions[c]-1]
  c=c+1
}

supplementary_table_5$library<-lib_code_to_name(supplementary_table_5$library, libraries, libraries_names)

supplementary_table_5<-supplementary_table_5[,.(library, assay, id, WT_aa, Pos, Pos_ref, Mut, ddg, std_ddg, ci95_ddg, structural_alignment_pos, binding_interface_contacts, binding_interface_5A, scHAmin_ligand, consv, secondary_structure, rsasa, structure_location)]
setnames(supplementary_table_5, "consv", "sequence_conservation")
fwrite(supplementary_table_5, "supplementary_tables/Supplementary_table5.txt")
```


# create medians ddg table
```{r}
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))
pdb_metrics_dt[,pdz:=sub("^([^_]+_[^_]+).*", "\\1", library)]

aggregated_data <- all_weights_data_complete[
  , .(median_ddg = median(ddg, na.rm = TRUE)),  # Calculate median of ddg
  by = .(library, structural_alignment_pos, Pos,assay, WT_aa, pdz)
]

all_weights_data_complete_median<-left_join(aggregated_data, pdb_metrics_dt, by=c("pdz", "Pos", "structural_alignment_pos", "WT_aa", "library"))

write.csv(all_weights_data_complete_median, paste0("tmp_tables/all_median_ddg_table.csv"))
```

# complete supplementary5 with the medians
```{r}
supplementary_table_5<-left_join(supplementary_table_5, all_weights_data_complete_median[,.(library, Pos, median_ddg)], by=c("library", "Pos"))

fwrite(supplementary_table_5, "supplementary_tables/Supplementary_table5.txt")
```


# median heatmaps
```{r, fig.width=20, fig.height=4}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))
gg_medians_alignment_heatmap(all_ddg_table)+geom_tile(data = all_ddg_table[binding_interface_contacts==T], fill = "transparent", color="green", size = 0.7)+theme(axis.text.x=element_text(angle=90))

gg_medians_alignment_heatmap(all_ddg_table[nchar(library)==nchar(pdz)], assay_plotted = "folding")+theme(axis.text.x=element_text(angle=90))
```

# complete ddG heatmaps
```{r}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_ddg_table.csv")))
all_ddg_table[,Mut:=substr(id, nchar(id), nchar(id))]
all_ddg_table$class<-all_ddg_table$structure_location
all_ddg_table[binding_interface_5A==T]$class<-"BI"
all_ddg_table[assay=="binding" & structural_alignment_pos %in% c(24,25,26)]$class<-"CBL"

pdb_metrics_dt$pdz<-as.character(pdb_metrics_dt$pdz)
all_ddg_table$pdz<-as.character(all_ddg_table$pdz)


for(lib in libraries_names){
  all_ddg_table_subset<-all_ddg_table[library==lib]
  pdz<-sub("^([^_]+_[^_]+).*", "\\1", lib)
  
  start_positions<-c(311, 311, 157, 157, 157, 123, 123, 1,1,1,1321,1321)
  all_ddg_table_subset[, Pos_ref:=0]
  all_ddg_table_subset$Pos_ref<-as.numeric(all_ddg_table_subset$Pos_ref)
  all_ddg_table_subset[, Pos_ref:=Pos+start_positions[which(libraries_names==lib)]-1]
  
  
  all_ddg_table_subset[,Pos:=Pos_ref]
  if(nchar(pdz)!=nchar(lib)){p<-gg_ddg_heatmap(all_ddg_table_subset[!is.na(Pos)], aa_order2, colors_aa, gsub("_", "-", gsub(" binding ", " | ", libraries_names[which(libraries==lib)])))}
  
  else{
    p<-gg_ddg_heatmap(all_ddg_table_subset[!is.na(Pos)], aa_order2, colors_aa, gsub("_", "-", gsub(" binding ", " | ", libraries_names_plots[which(libraries_names==lib)])))+
  theme(legend.position = "right")
  }
    
  p+theme(legend.position="none", axis.text.x=element_text(size=22, face="bold"),axis.text.y=element_text(size=20, face="bold"))
  ggsave(paste0("Figs/Fig2/ddg_heatmaps/ddG_", lib, "_heatmap.png"), width=18, height=6)
}
```

# structure shapes
```{r}
all_median_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))

for(lib in libraries_names){
  pdz<-sub("^([^_]+_[^_]+).*", "\\1", lib)
  
  if(pdz %in% c("psd95_pdz2", "erbin_pdz1")){point_height=1.05}
  if(pdz %in% c("psd95_pdz3")){point_height=1.03}
  if(pdz %in% c("nherf3_pdz2")){point_height=1.11}
  if(pdz %in% c("nherf3_pdz1")){point_height=1.09}
  
  ddg_subset<-all_median_ddg_table[library==lib]
  
  p<-ggplot(ddg_subset,aes(x=as.factor(Pos), y=factor(library)))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=median_ddg), color="black")+ 
    theme_void(base_size = 16)+
    scale_x_discrete(limits = seq(1, max(ddg_subset$Pos, na.rm = T), 1))+
    scale_fill_gradient2(limits=c(-0.8932919, 3.670025),"median ∆∆G",high="red",mid="grey90",low="blue",midpoint=0, na.value = "white")+
    theme( plot.title = element_text(size=20))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           legend.position = "none")+
    ggtitle(paste0(""))+
    geom_point(data=ddg_subset[binding_interface_contacts==T], aes(y=point_height), shape=21, fill="red", size=2) 
  
  
  ggsave(paste0("Figs/Fig2/colored_heatmap/", lib, ".png"), width=10, height=1.5)
}
```
Create the shaped heatmap
```{r}
library(magick)

for(lib in libraries_names){
  print(lib)
  path_image<-paste0("Figs/Fig2/colored_heatmap/", lib,".png")
  path_shape<-paste0("data/shapes/", lib, "_shape.png")
  out_path<-paste0("Figs/Fig2/colored_shapes/", lib, "_median_ddg.png")
  crop_heatmap_structure_shape_outline(path_image, path_shape, out_path)
}
```


# load some tables
```{r}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_ddg_table.csv")))

#preparing some data
all_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

all_median_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_median_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

all_ddg_table[, Mut:=substr(id, nchar(id), nchar(id))]

#prepare binding tables
all_ddg_table_binding<-all_ddg_table[assay=="binding"]
all_ddg_table_folding<-all_ddg_table[assay=="folding"]
all_median_ddg_table_binding<-all_median_ddg_table[assay=="binding"]

#list all + common binding interface positions
all_bi<-unique(all_ddg_table_binding[ binding_interface_contacts==T]$structural_alignment_pos)
core_bi<-unique(all_ddg_table_binding[core_bi_contacts==T]$structural_alignment_pos)

#classify these positions
all_ddg_table_binding[,class_BI:="outside_BI"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi,class_BI:="some_BI, contact"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi & binding_interface_contacts==F,class_BI:="some_BI, NOcontact"]
all_ddg_table_binding[structural_alignment_pos %in% core_bi,class_BI:="core_BI"]
```

# correlations of all effects (figure 2)
```{r}
library(data.table)
library(ggplot2)

#  Prepare and simplify
dt_b <- all_ddg_table_binding[assay=="binding", .(library, id, Mut, structural_alignment_pos, ddg_b = ddg)]
dt_f <- all_ddg_table_folding[assay=="folding", .(library, id, Mut, structural_alignment_pos, ddg_f = ddg)]

#  Merge both for reference positions/mutations
dt_all <- merge(dt_b, dt_f, by = c("library", "structural_alignment_pos", "Mut"), all = TRUE)


calc_ddg_cors <- function(dt, ddg_colname = "ddg_b") {
  libs <- unique(dt$library)
  lib_pairs <- combn(libs, 2, simplify = FALSE)
  
  rbindlist(lapply(lib_pairs, function(pair) {
    lib1 <- pair[1]; lib2 <- pair[2]
    
    d1 <- dt[library == lib1, .(structural_alignment_pos, id,Mut, ddg1 = get(ddg_colname))]
    d2 <- dt[library == lib2, .(structural_alignment_pos, id,Mut, ddg2 = get(ddg_colname))]
    
    merged <- merge(d1, d2, by = c("structural_alignment_pos", "Mut"))
    
    cor_val <- if (nrow(merged) > 2) {
      cor(merged$ddg1, merged$ddg2, use = "pairwise.complete.obs", method = "pearson")
    } else NA_real_
    
    data.table(lib1 = lib1, lib2 = lib2, correlation = cor_val)
  }))
}

# Compute correlations for binding and folding
cors_b <- calc_ddg_cors(dt_b, "ddg_b")[, assay := "binding"]
cors_f <- calc_ddg_cors(dt_f, "ddg_f")[, assay := "folding"]


cors_all <- rbind(cors_b, cors_f, fill = TRUE)


# Split lib1 and lib2 into their components: gene, pdz, ligand
cors_all[, c("gene1", "pdz1", "ligand1") := tstrsplit(lib1, "_", fixed = TRUE)]
cors_all[, c("gene2", "pdz2", "ligand2") := tstrsplit(lib2, "_", fixed = TRUE)]


cors_all[, PDZ1 := paste(gene1, pdz1, sep = "_")]
cors_all[, PDZ2 := paste(gene2, pdz2, sep = "_")]
cors_all[, class := "unrelated"]
cors_all[PDZ1==PDZ2, class := "same_pdz"]
cors_all[ligand1==ligand2, class := "same_ligand"]
cors_all[lib1==lib2, class := "same_pdz"]

dt_plot <- cors_all[!is.na(correlation)]

counts_dt <- dt_plot[
  , .(n = .N),
  by = .(class, assay)
]

# Plot distributions
ggplot(cors_all[!is.na(correlation)], aes(x=factor(class, levels=c("same_pdz", "same_ligand", "unrelated"),labels=c("Same\nPDZ", "Same\nligand", "Unrelated")), y=correlation, color=class, fill=class))+
  geom_boxplot(alpha=0.5)+
  geom_point(position="jitter", alpha=0.5)+
  ylab("pairwise pearson correlation")+
  theme_classic()+
  theme(axis.title.x=element_blank(), legend.position="none")+
  scale_color_manual(values=c( "#9B7EBD", "#B0DB9C", "#DA6C6C"))+
  scale_fill_manual(values=c( "#9B7EBD", "#B0DB9C", "#DA6C6C"))+
  geom_text(
    data = counts_dt,
    aes(
      x = factor(class,
                 levels = c("same_pdz", "same_ligand", "unrelated"),
                 labels = c("Same\nPDZ", "Same\nligand", "Unrelated")),
      y = 0.02,
      label = paste0("n = ", n)
    ),
    inherit.aes = FALSE,
    size = 3.5,
    fontface = "bold"
  )+
  ylim(c(0,1))+theme(axis.text.x=element_text(size=9,face="bold", color="black"), 
        axis.text.y=element_text(size=13,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))+
  facet_grid(~assay, scales="free_x", space="free_x")




ggsave(paste0("Figs/Fig2/Fig2H_General_correlations_ddg.pdf"), width=6, height=4)
```


# F vs B with coloured groups
```{r}
all_plots<-list()
counter=1
all_ddg_table_f_vs_b<-data.table()
for(lib in libraries_binding_names){
  all_ddg_table_b<-all_ddg_table[library==lib]
  pdz<-unique(all_ddg_table_b$pdz)
  all_ddg_table_f<-all_ddg_table[library==pdz]
  all_ddg_table_subset<-merge(all_ddg_table_f[,c("id","ddg")],all_ddg_table_b[,c("id","ddg", "binding_interface_contacts","structure_location")], by="id", suffix=c("_f", "_b"))
  all_ddg_table_subset[,class:=structure_location]
  all_ddg_table_subset[binding_interface_contacts==T,class:="binding_interface"]
  
  all_ddg_table_subset[, library:=lib]
  all_ddg_table_f_vs_b<-rbind(all_ddg_table_f_vs_b, all_ddg_table_subset)
  counter=counter+1
}

libraries_names_<-libraries_names
libraries_names_[libraries_names_=="nherf3_pdz1_dap-1"]<-"nherf3_pdz1_dap_1"
all_ddg_table_f_vs_b$library<-lib_code_to_name(gsub("-", "_",all_ddg_table_f_vs_b$library), libraries_names_, libraries_names_plots)
all_ddg_table_f_vs_b$library<-gsub("_", "-", gsub(" binding ", "\n", all_ddg_table_f_vs_b$library))
ggplot(all_ddg_table_f_vs_b, aes(x=ddg_f, y=ddg_b, color=class))+
  geom_hline(aes(yintercept=0), color="grey")+
  geom_vline(aes(xintercept=0), color="grey")+
    geom_point(data=all_ddg_table_f_vs_b[class!="binding_interface",],alpha=0.6, size=0.5)+
    geom_point(data=all_ddg_table_f_vs_b[binding_interface_contacts==T,], color="red",alpha=0.6, size=0.5)+
    scale_color_manual(values=c("green3", "blue"))+
    facet_wrap(~library, nrow=1)+
    theme_classic()+
    ylim(c(-2,3))+
    #xlim(c(-1,2.5))+#ggtitle(gsub(" binding ", " | ", gsub( "_", "-",libraries_names[which(libraries==lib)])))+
    xlab("∆∆Gf")+
    ylab("∆∆Gb")+theme(title=element_text(size=7, face="bold"),
                       axis.title.x=element_text(size=15, face="bold"),
                       axis.title.y=element_text(size=15, face="bold"), legend.position = "none")
ggsave(paste0("Figs/Fig2/FigS2Eddg_folding_vs_binding.pdf"), width=10, height=2.5)
```


## example
```{r}
dt<-all_ddg_table_f_vs_b[library=="ERBIN-PDZ1\nCRIPT"]
ggplot(dt, aes(x=ddg_f, y=ddg_b, color=class))+
  geom_hline(aes(yintercept=0), color="grey")+
  geom_vline(aes(xintercept=0), color="grey")+
    geom_point(data=dt[class!="binding_interface",],alpha=0.6, size=0.5)+
    geom_point(data=dt[binding_interface_contacts==T,], color="red",alpha=0.6, size=0.5)+
    scale_color_manual(values=c("green3", "blue"))+
    facet_wrap(~library, nrow=1)+
    theme_classic()+
    ylim(c(-2,3))+
    xlab("∆∆Gf")+
    ylab("∆∆Gb")+theme(title=element_text(size=7, face="bold"),
                       axis.title.x=element_text(size=15, face="bold"),
                       axis.title.y=element_text(size=15, face="bold"), legend.position = "none")
ggsave(paste0("Figs/Fig2/Fig2Iddg_folding_vs_binding.pdf"), width=3, height=4)
```


# core vs surface
```{r}

all_ddg_table$lib_name<-lib_code_to_name(all_ddg_table$library, libraries_names, libraries_names_plots)
all_ddg_table$lib_name<-gsub(" a", "\na", all_ddg_table$lib_name)

all_ddg_table$lib_name<-gsub("_", "-", all_ddg_table$lib_name)


p_dt <- all_ddg_table[assay=="folding", .(p = wilcox.test(ddg ~ structure_location)$p.value), by = lib_name]
p_dt
p_dt[, label := sprintf("p = %.2e", p)]
p_dt[, `:=`(group1 = "core",group2 = "surface", xmin = "core",xmax = "surface",y.position = 4.3)]



N<-all_ddg_table[assay=="folding", .(.N), by=.(structure_location, lib_name)]
ggplot(all_ddg_table[assay=="folding"], aes(x=structure_location, y=ddg, color=structure_location))+
  facet_wrap(~lib_name, ncol=2)+
  ylab("∆∆Gf")+
  geom_violin(aes(fill=structure_location))+
  geom_boxplot(width=0.1, color="black", outliers = F)+
  theme_classic()+
  #stat_compare_means(comparisons=list(c("core", "surface")),aes(label = after_stat(p.signif)), method="wilcox.test")+
  theme(axis.title.x=element_blank())+
  scale_color_manual(values=c("green3", "blue"))+
  scale_fill_manual(values=c("green3", "blue"))+
  ylim(c(-2.5,5.5))+theme(legend.position="none")+
  theme(title=element_text(size=7, face="bold"),
                       axis.title.x=element_blank(),
        axis.text.x=element_text(size=10, face="bold"),
                       axis.title.y=element_text(size=15, face="bold"), legend.position = "none")+
   geom_text(data=N, aes(label=paste0("N=", N), y=-2), color="black", face="bold", size=3)+
  stat_pvalue_manual(p_dt, label = "label", xmin = "xmin",xmax = "xmax",y.position = "y.position",tip.length = 0.01)

ggsave(paste0("Figs/Fig2/FigS2Ddistribution_ddg_core_surface.pdf"), width=3, height=7)
```


```{r}
dt<-all_ddg_table[assay=="folding" & pdz =="erbin_pdz1"]
N<-dt[, .(.N), by=structure_location]
N$lib_name<-"ERBIN-PDZ1\nabundance"

p_dt <- dt[, .(p = wilcox.test(ddg ~ structure_location)$p.value), by = lib_name]
p_dt
p_dt[, label := sprintf("p = %.2e", p)]
p_dt[, `:=`(group1 = "core",group2 = "surface", xmin = "core",xmax = "surface",y.position = 5)]

ggplot(dt, aes(x=structure_location, y=ddg, color=structure_location))+
  facet_wrap(~lib_name, ncol=2)+
  ylab("∆∆Gf")+
  geom_violin(aes(fill=structure_location))+
  geom_boxplot(width=0.1, color="black", outliers = F)+
  theme_classic()+
  #stat_compare_means(comparisons=list(c("core", "surface")), label = NULL, hide.ns = TRUE, method="wilcox.test")+#aes(label = NA)
  theme(axis.title.x=element_blank())+
  scale_color_manual(values=c("green3", "blue"))+
  scale_fill_manual(values=c("green3", "blue"))+
  ylim(c(-2.5,5.5))+theme(legend.position="none")+
  theme(title=element_text(size=7, face="bold"),
                       axis.title.x=element_blank(),
        axis.text.x=element_text(size=10, face="bold"),
                       axis.title.y=element_text(size=15, face="bold"), legend.position = "none")+
  geom_text(data=N, aes(label=paste0("N=", N), y=-2), color="black", face="bold")+
  stat_pvalue_manual(p_dt, label = "label", xmin = "xmin",xmax = "xmax",y.position = "y.position",tip.length = 0.01)

ggsave(paste0("Figs/Fig2/Fig2Jdistribution_ddg_core_surface.pdf"), width=2.5, height=4)
```

