---
title: "PDZ Binding interfaces"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---


# initializing
```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

# load some variables
load('source_variables.RData')
```

## libraries used for the analysis
```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(bio3d)
library(dplyr)
library(stringr)
library(readxl)
library(tidyr)
library(cowplot)
library(ggpmisc)
library(ggpubr)
library(scales)
library(ggridges)
library(ggiraphExtra)

sessionInfo()

```


# load some tables
```{r}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_ddg_table.csv")))
all_median_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))
aln_table<-as.data.table(read.table(paste0("data/annotations/structural_alignment_tcoffee.csv")))
get_contacts_dt<-as.data.table(read.csv("data/annotations/tmp_files/contacts_table_complete.csv", sep=","))
```


# preparing some variables
```{r}
#preparing some data
all_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

all_median_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_median_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

all_ddg_table[, Mut:=substr(id, nchar(id), nchar(id))]

#prepare binding tables
all_ddg_table_binding<-all_ddg_table[assay=="binding"]
all_ddg_table_folding<-all_ddg_table[assay=="folding"]
all_median_ddg_table_binding<-all_median_ddg_table[assay=="binding"]

#list all + common binding interface positions
all_bi<-unique(all_ddg_table_binding[ binding_interface_contacts==T]$structural_alignment_pos)
core_bi<-unique(all_ddg_table_binding[core_bi_contacts==T]$structural_alignment_pos)

# saving this lists
save(all_bi, file = paste0("tmp_tables/all_bi.Rdata"))
save(core_bi, file = paste0("tmp_tables/core_bi.Rdata"))

#classify these positions
all_ddg_table_binding[,class_BI:="outside_BI"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi,class_BI:="some_BI, contact"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi & binding_interface_contacts==F,class_BI:="some_BI, NOcontact"]
all_ddg_table_binding[structural_alignment_pos %in% core_bi,class_BI:="core_BI"]
```



# BI distribution of effects

```{r}
all_ddg_table_binding$lib_name<-lib_code_to_name(all_ddg_table_binding$library, libraries_binding_names, libraries_names_plots[which(libraries %in% libraries_binding)])
all_ddg_table_binding$lib_name<-gsub(" binding ", "\n", all_ddg_table_binding$lib_name)
all_ddg_table_binding$lib_name<-gsub("_", "-", all_ddg_table_binding$lib_name)

p<-ggplot(all_ddg_table_binding[binding_interface_contacts==T], aes(x=ddg))+
  theme_classic()+
  xlab("∆∆Gb in binding\ninterface positions")+
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x=element_text(angle=90, vjust=0.5, hjust = 1, size=7))+
  xlim(c(-1.5,3))+
  geom_density_ridges2(aes(y=lib_name), scale = 1.2, alpha=0.8, fill="grey80", color="grey50")+coord_flip()+
  theme(axis.text.x=element_text(size=9,face="bold", color="black"), 
        axis.text.y=element_text(size=13,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))+
  geom_hline(aes(yintercept=0))

ggsave(paste0("Figs/Fig3/Fig3AddGb_distribution_BI.pdf"), width=6, height=3)
```


## Identify hotspots
Defining hotspots as residues where mutations cause changes in binding energy larger than the median of all interface residues (0.75 kcal/mol)
```{r, fig.width=10, fig.height=10}
# all candidate positions
plot_dt<-all_ddg_table_binding[binding_interface_contacts==T]
summary(plot_dt$ddg,na.rm = T)

# Calculate the median of all BI residues
median_BI <- median(all_ddg_table_binding[binding_interface_contacts == T]$ddg, na.rm = TRUE)

# Filter groups with at least 2 observations
filtered_data <- plot_dt[, 
  .N, by = .(library, structural_alignment_pos)][N >= 2]

# Merge the filtered groups back with the original data
valid_data <- plot_dt[library %in% filtered_data$library &  structural_alignment_pos %in% filtered_data$structural_alignment_pos,]


# Perform the t-test only for valid groups
test_results <- valid_data[
  !is.na(ddg),  # Ensure ddg is not NA
  .(
    p_value = if (.N > 1 && length(unique(ddg)) > 1) {
      t.test(ddg, mu = median_BI, alternative = "greater")$p.value
    } else {
      NA_real_  # Use numeric NA for consistency
    }
  ),
  by = .(library, structural_alignment_pos)
]

test_results[, p_adjusted := p.adjust(p_value, method = "fdr")]
significant_distributions<-test_results[p_adjusted<0.1] # save the significant ones in a new distribution
significant_distributions[,significant:=T]
plot_dt<-left_join(plot_dt, significant_distributions)

# CONSERVED binding interface hotspots
# define the conserved binding interface hotspots those with >=4 libraries significantly above the median
BI_hotspots<-significant_distributions[significant==T, .N, by=c("structural_alignment_pos")][N>=4]$structural_alignment_pos
save(BI_hotspots, file = paste0("tmp_tables/BI_hotspots.Rdata"))

# Add a new column to label significant facets
all_ddg_table_binding[, highlight := ifelse(structural_alignment_pos %in% BI_hotspots, "highlight", "normal")]

plot_dt[is.na(significant), significant:=F]
plot_dt[, BI_consv_hotspot:=(structural_alignment_pos %in% BI_hotspots)]
BI_hotspots_dt<-plot_dt
fwrite(BI_hotspots_dt, paste0("tmp_tables/BI_hotspots_dt.csv"))


# add also the 
plot_dt2<-merge(plot_dt, all_ddg_table[assay=="binding" & structural_alignment_pos %in% plot_dt$structural_alignment_pos, c("library", "id", "structural_alignment_pos", "ddg")], by=c("id", "structural_alignment_pos", "library"), all =T)
setnames(plot_dt2, "ddg.y", "ddg")

# prepare library names annotations
plot_dt2$lib_name<-lib_code_to_name(plot_dt2$library, libraries_binding_names, libraries_names_plots[which(libraries %in% libraries_binding)])
plot_dt2$lib_name<-gsub(" binding ", " | ", plot_dt2$lib_name)
plot_dt2$lib_name<-gsub("_", "-", plot_dt2$lib_name)

# annotate secondary structure labels
plot_dt2[, region:=""]
plot_dt2[structural_alignment_pos %in% seq(24,30), region:="β2"]
plot_dt2[structural_alignment_pos %in% c(24,25,26), region:="carboxylate binding loop"]
plot_dt2[structural_alignment_pos %in% seq(30,45), region:="β2-β3 loop"]
plot_dt2[structural_alignment_pos %in% seq(46,52), region:="β3"]
plot_dt2[structural_alignment_pos %in% seq(82,91), region:="α3"]
plot_dt2[, region:=structural_alignment_pos]

plot_dt2$significant<-as.character(plot_dt2$significant)
plot_dt2[is.na(significant), significant:="non_contact"]

n_boxplots<-plot_dt2[, .(.N), by=.(region, structural_alignment_pos, lib_name)]

library(ggnewscale)
p<-ggplot(plot_dt2)+
  geom_hline(aes(yintercept = median(plot_dt$ddg,na.rm = T)), color="red3")+
  geom_hline(aes(yintercept = 0), color="grey", linetype="dashed")+
  geom_boxplot(aes(x=factor(structural_alignment_pos), y=ddg, fill=lib_name, color=significant, alpha=significant, linewidth = significant))+
  scale_fill_brewer(palette="Set2")+
  theme_pubr()+
  scale_color_manual("Fdr<0.1",values=c("grey60", "grey90", "black"))+
  scale_linewidth_manual(values=c(0.4, 0.3, 0.7))+
  scale_alpha_manual(values=c(1,0.2,1))+
  facet_wrap(~region+factor(structural_alignment_pos), scales="free_x", ncol=7)+#, ncol=13)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x =element_blank(),
        legend.position="right",
    strip.text = element_text(
      color="black",
      size = 10,
      face = "bold"
    )
  ) +ylab("∆∆G")+xlab("")


p+theme(legend.position="none")
ggsave(paste0("Figs/Fig3/Fig3Bdefine_hotspots.pdf"), width=10, height=10)

fwrite(plot_dt, paste0("tmp_tables/BI_hotspots_dt.csv"))

```

# number of points in each boxplot
```{r}
n_boxplots
```



# percentage of hotspots per interaction
```{r}
BI_hotspots_dt_<-BI_hotspots_dt[, .(significant=significant[1]), by=.(structural_alignment_pos, library)]
BI_hotspots_dt_[,pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_counts_<-BI_hotspots_dt_[, .(.N, N_sign=sum(significant)), by=.(library)]
BI_hotspots_dt_counts_[,percentage:=N_sign/N*100]
summary(BI_hotspots_dt_counts_$percentage)
```




```{r}
BI_hotspots_dt_<-BI_hotspots_dt[, .(significant=significant[1]), by=.(structural_alignment_pos, library)]
BI_hotspots_dt_[,pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_libraries<-BI_hotspots_dt_[, .(N=.N, num_hotspots=sum(significant)), by=.(structural_alignment_pos)]
BI_hotspots_dt_pdz<-BI_hotspots_dt_[significant==T, .(num_pdzs=length(unique(pdz))), by=.(structural_alignment_pos)]


BI_hotspots_dt_pdz_nohotspots<-all_bi[ !all_bi%in% unique(BI_hotspots_dt_pdz$structural_alignment_pos)]

BI_hotspots_dt_libraries[num_hotspots==7] # which are hotspots in all interactions
BI_hotspots_dt_pdz[num_pdzs==5] # which are hotspots in all 5 PDZs
BI_hotspots_dt_pdz[num_pdzs>=3] # which are hotspots in half of the PDZs
```


```{r}
#never classified as hotspots
BI_hotspots_dt_[, .(num_pdzs=sum(significant)), by=.(structural_alignment_pos)][num_pdzs==0]

# never classified as hotspots when interacting in several PDZs
BI_hotspots_dt_[, pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_[, .(num_pdzs=sum(significant), total_contacts_pdzs=length(unique(pdz))), by=.(structural_alignment_pos)][num_pdzs==0 & total_contacts_pdzs>1][order(structural_alignment_pos)]
```



```{r warning=FALSE, fig.width=3, fig.height=3}
BI_hotspots_dt_none_sign<-BI_hotspots_dt_[, .(num_pdzs=sum(significant)), by=.(structural_alignment_pos)][num_pdzs==0]
BI_hotspots_dt_pdz<-rbind(BI_hotspots_dt_pdz, BI_hotspots_dt_none_sign)
BI_hotspots_dt_pdz<-rbind(BI_hotspots_dt_pdz, data.table(structural_alignment_pos=0, num_pdzs=2))

ggplot(BI_hotspots_dt_pdz)+
  geom_bar(aes(x=factor(num_pdzs, levels=c(0,1,2,3,4,5)), fill=factor(num_pdzs)), stat="count")+
  scale_fill_manual(values=c("grey", "black","white", "brown1","brown1", "brown1", "brown1", "brown1"))+
  ylab("Position count")+
  xlab("# PDZs with hotspot")+
  theme_classic()+theme(legend.position="none")+
  theme(axis.text.x=element_text(size=10,face="bold", color="black"), 
        axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"),
        axis.title.x=element_text(size=10,face="bold", color="black"))
ggsave(paste0("Figs/Fig3/Fig3Clibraries_per_hotspot.pdf"), width=2.5, height=2.5)
```


# Position contacts of hotspots (histogram)
```{r}
get_contacts_dt_aln_ligand<-get_contacts_dt[,Pos:=Pos_1][chain_2=="B" & chain_1=="A", .(lig_contacts=paste(Pos_2-9, collapse=",")), by=.(Pos, library)]

BI_hotspots_dt_w_contacts<-left_join(get_contacts_dt_aln_ligand, BI_hotspots_dt[,.(significant=significant[1]) ,by=.(library, Pos, structural_alignment_pos)][,.(library, Pos, structural_alignment_pos, significant)])

# all of the conserved residues conract position 0 and/or -2
BI_hotspots_dt_w_contacts[significant==T & structural_alignment_pos %in% BI_hotspots]

BI_hotspots_dt_w_contacts[, has_0_or_neg2 := grepl("(^|,)(-2|0)(,|$)", lig_contacts)]
BI_hotspots_dt_w_contacts[, conserved_hotspot:=structural_alignment_pos %in% BI_hotspots]

#enrichment of has_0_or_neg2 in overall hotspots
fisher.test(table(BI_hotspots_dt_w_contacts$has_0_or_neg2, BI_hotspots_dt_w_contacts$significant), alternative = "greater" )
```


```{r}
# plot the number of contacts
BI_hotspots_dt_w_contacts[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt_w_contacts[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt_w_contacts[conserved_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]

dt_expanded <- BI_hotspots_dt_w_contacts[, .(lig_pos = unlist(strsplit(lig_contacts, ","))), by = .(hotspot_class)]
ligand_contact_counts <- dt_expanded[, .N, by = .(hotspot_class, lig_pos)][order(hotspot_class, lig_pos)]

ggplot(ligand_contact_counts, aes(x=factor(as.numeric(lig_pos)), y=N, fill=hotspot_class))+
  geom_bar(stat="identity")+
  facet_wrap(~factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), ncol=1)+
  scale_x_discrete(limits = factor(c(0:-8)))+xlab("ligand position")+ylab("number of contacts")+
  theme_classic()+
  theme(legend.position = "none", strip.text = element_blank())+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  theme(axis.text.x=element_text(size=10,face="bold", color="black"), 
        axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"),
        axis.title.x=element_text(size=10,face="bold", color="black"))
ggsave(paste0("Figs/Fig3/Fig3Ehotspot_contacts.pdf"), width=3, height=3)
```


# conserved hotapots are highly conserved structural contacts
```{r}
#enrichment in core_bi residues
dt_tmp<-data.table(all_bi=all_bi)
dt_tmp[all_bi %in% BI_hotspots, BI_hotspot:=T]
dt_tmp[!all_bi %in% BI_hotspots, BI_hotspot:=F]
dt_tmp[all_bi %in% core_bi, core_BI:=T]
dt_tmp[!all_bi %in% core_bi, core_BI:=F]
fisher.test(table(dt_tmp$core_BI, dt_tmp$BI_hotspot), alternative = "greater")
```


# conservarion in sequence
```{r, fig.width=4, fig.height=4}
BI_hotspots_dt_tmp<-BI_hotspots_dt
BI_hotspots_dt_tmp[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt_tmp[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt_tmp[BI_consv_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple<-BI_hotspots_dt_tmp[, .(significant=significant[1], consv=consv[1], hotspot_class=hotspot_class[1]), by=.(library, structural_alignment_pos)]

BI_hotspots_dt_simple<-BI_hotspots_dt_simple[,.(hotspot_class=hotspot_class[1], consv=consv[1]), by=.(structural_alignment_pos)]
BI_hotspots_dt_simple[structural_alignment_pos %in% BI_hotspots_dt_simple, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple[structural_alignment_pos %in% BI_hotspots_dt_none_sign$structural_alignment_pos, hotspot_class:="non-Hotspot"]

n_boxplots<-BI_hotspots_dt_simple[, .(.N), by=hotspot_class]

ggplot(BI_hotspots_dt_simple, aes(x=factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), y=consv, fill=hotspot_class, color=hotspot_class))+
  geom_violin(scale="width", alpha=0.7)+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  scale_color_manual(values=c("black", "brown1", "grey"))+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  theme_classic()+
  ylab("Sequence conservation score")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), axis.title.x=element_blank(), legend.position="none")+
  stat_compare_means(comparisons = list(c("Hotspot in >2/5 PDZs", "Hotspot"), c("Hotspot", "non-Hotspot")), aes(label = after_stat(p)),method = "wilcox.test")+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))+
  geom_text(data=n_boxplots, aes(x=hotspot_class, y=-0.5, label=paste0("N=", N)), color="black")

ggsave(paste0("Figs/Fig3/Fig3Fseq_consv_hotspots.pdf"), width=2.3, height=2.3)
```



# conservation in structure
previously superimposed the structures using Matchmaker from ChimeraX,  
And then manually computed the RMSD of the residue subsets of interest
```{r warning=FALSE, fig.width=4, fig.height=4}
#load models
library(bio3d)
model1 <- read.pdb(paste0("data/structures/superimposed/superimposed_dlg4_pdz3_cript.pdb.pdb"))
model2 <- read.pdb(paste0("data/structures/superimposed/superimposed_dlg4_pdz2_cript.pdb.pdb"))
model3 <- read.pdb(paste0( "data/structures/superimposed/superimposed_dlg4_pdz2_nmdar2a.pdb.pdb"))
model4 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz2_cep164.pdb.pdb"))
model5 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz1_dlgap.pdb.pdb"))
model6 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz1_uhmk1.pdb.pdb"))
model7 <- read.pdb(paste0("data/structures/superimposed/superimposed_erbin_pdz1_cript.pdb.pdb"))
models<-list(model1, model2, model3, model4, model5, model6, model7)


#define the function that will compute the local rmsd (actually local euclidean distances)
calculate_local_rmsd <- function(model1, model2, pair) {
  index1<-model1$atom$resno==pair[1] &  model1$atom$elety=="CA" & model1$atom$chain=="A"
  index2<-model2$atom$resno==pair[2] &  model2$atom$elety=="CA" & model2$atom$chain=="A"
  coord1 <- c(model1$atom$x[index1], model1$atom$y[index1], model1$atom$z[index1])
  coord2 <- c(model2$atom$x[index2], model2$atom$y[index2], model2$atom$z[index2])
   # Euclidean distance (squared differences summed)
  rmsd <- sqrt(sum((coord1 - coord2)^2))   # Root Mean Square Deviation
  return(rmsd)
}

# build the local (euclidean) distances between aligned residue's CA atoms
combinations<-t(combn(1:7, 2))
rmsd_comparisons<-data.table()
for(i in 1:nrow(combinations)){
  pair<-combinations[i,]
  libraries_<-libraries_binding_names[pair]
  model1<-models[pair[1]]
  model2<-models[pair[2]]
  
  #paired positions
  paired_dt<-t(reshape(all_median_ddg_table_binding[library %in% libraries_, c("structural_alignment_pos", "Pos", "library")], timevar="structural_alignment_pos", idvar="library", direction="wide"))
  aln_positions<-as.numeric(lapply(strsplit(rownames(paired_dt[-1,]), "[.]"), "[[", 2))
  paired_dt<-data.table(paired_dt[-1,])
  paired_dt$structural_alignment_pos<-aln_positions
  paired_dt <- na.omit(paired_dt)
  
  for(b_a_p in paired_dt$structural_alignment_pos){
    positions<-as.numeric(c(paired_dt[structural_alignment_pos==b_a_p]$V1,paired_dt[structural_alignment_pos==b_a_p]$V2))
    local_rmsd <- calculate_local_rmsd(model1[[1]], model2[[1]], positions)
    rmsd_comparisons<-rbind(rmsd_comparisons, data.table( paste0(libraries_, collapse="."), libraries_[1], libraries_[2], local_rmsd, b_a_p))
  }

}

colnames(rmsd_comparisons)<-c("lib_pair", "lib1", "lib2", "local_rmsd", "structural_alignment_pos")
rmsd_comparisons<-left_join(rmsd_comparisons[,library:=lib1], BI_hotspots_dt[,.(library, structural_alignment_pos, significant)])[, hotspot1:=significant]
rmsd_comparisons<-left_join(rmsd_comparisons[,library:=lib2], BI_hotspots_dt[,.(library, structural_alignment_pos, significant)])[, hotspot2:=significant]

rmsd_comparisons_studyBI<-rmsd_comparisons
rmsd_comparisons_studyBI[is.na(hotspot1) & is.na(hotspot2), hotspot:="non-hotspot"]
rmsd_comparisons_studyBI[!is.na(hotspot1) | !is.na(hotspot2), hotspot:="hotspot"]
rmsd_comparisons_studyBI[!is.na(hotspot1) & !is.na(hotspot2), hotspot:="both hotspot"]
rmsd_comparisons_studyBI[structural_alignment_pos %in% BI_hotspots & hotspot!="non-hotspot" & hotspot!="hotspot", hotspot:="conserved hotspot"]


rmsd_results <- rmsd_comparisons_studyBI[, .(
  non_hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "non-hotspot"])^2)) else NA_real_,
  consv_hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "conserved hotspot"])^2)) else NA_real_,
  hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "both hotspot"])^2)) else NA_real_
), by=lib_pair]

# annotate libraries
dt_tmp<-melt(rmsd_results, id.var="lib_pair")
dt_tmp[, class2 := as.integer(grepl("4", lib_pair))]

dt_tmp[, lib1 := sub("\\..*", "", lib_pair)]      # before the first dot
dt_tmp[, lib2 := sub("^[^.]*\\.", "", lib_pair)]  # after the first dot

dt_tmp[sub("^(([^_]+)_([^_]+))_.*", "\\1", lib1)!=sub("^(([^_]+)_([^_]+))_.*", "\\1", lib2), diff_pdz:="different PDZ"] #keep only comparisons between different PDZs
dt_tmp[sub("^(([^_]+)_([^_]+))_.*", "\\1", lib1)==sub("^(([^_]+)_([^_]+))_.*", "\\1", lib2), diff_pdz:="same PDZ"]


# plot the distributions per group]
n_boxplots<-dt_tmp[diff_pdz=="different PDZ"][, .(.N), by=variable]
ggplot(dt_tmp[diff_pdz=="different PDZ"], aes(x=factor(variable, levels=c("consv_hotspot","hotspot", "non_hotspot"), labels=c("BI conserved\nhotspot","hotspot", "non-hotspot")), y=value, fill=variable, color=variable))+
  geom_violin(scale="width", alpha=0.7)+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  stat_compare_means(comparisons=list(c("BI conserved\nhotspot","hotspot"), c("non-hotspot","hotspot")), size=5, aes(label=after_stat(p)), paired=T,method = "wilcox.test")+
  scale_fill_manual(values=c("grey","brown1", "black"))+
  scale_color_manual(values=c("grey","brown1", "black"))+
  theme_classic()+
  ylim(c(0,6.8))+
  ylab("RMSD")+
  theme(legend.position = "none", axis.title.x=element_blank())+
  theme(legend.position="none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))+
  geom_text(data=n_boxplots, aes(x=factor(variable, levels=c("consv_hotspot","hotspot", "non_hotspot"), labels=c("BI conserved\nhotspot","hotspot", "non-hotspot")), y=0, label=paste0("N=", N)), color="black")

ggsave(paste0("Figs/Fig3/Fig3Gstructural_conservation_hotspot.pdf"), width=2.5, height=2.5)

```

# rsasa of the hotspots

```{r, fig.width=4, fig.height=4}
BI_hotspots_dt[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt[BI_consv_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple<-BI_hotspots_dt[, .(significant=significant[1], rsasa=rsasa[1], hotspot_class=hotspot_class[1]), by=.(library, structural_alignment_pos)]

n_boxplots<-BI_hotspots_dt_simple[, .(.N), by=hotspot_class]
ggplot(BI_hotspots_dt_simple, aes(x=factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), y=rsasa, fill=hotspot_class, color=hotspot_class))+
  geom_hline(aes(yintercept = 0.25), color="grey")+
  geom_violin(scale="width", alpha=0.7)+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  scale_color_manual(values=c("black", "brown1", "grey"))+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  theme_classic()+
  ylab("RSASA")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), axis.title.x=element_blank(), legend.position="none")+
  stat_compare_means(comparisons = list(c("Hotspot in >2/5 PDZs", "Hotspot"), c("Hotspot in >2/5 PDZs", "non-Hotspot"), c("Hotspot", "non-Hotspot")),method = "wilcox.test", aes(label = after_stat(p)))+ylim(c(0,1.4))+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))+
  geom_text(data=n_boxplots, aes(x=factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), y=0, label=paste0("N=", N)), color="black")

ggsave(paste0("Figs/Fig3/Fig3Hrsasa_hotspots.pdf"), width=2.5, height=2.5)

summary(BI_hotspots_dt_simple[hotspot_class=="Hotspot in >2/5 PDZs"]$rsasa)
summary(BI_hotspots_dt_simple[hotspot_class=="Hotspot"]$rsasa)
summary(BI_hotspots_dt_simple[hotspot_class=="non-Hotspot"]$rsasa)
```





# chimera hotspots figures
```{r}
BI_hotspots_dt_<-BI_hotspots_dt[,.(significant=significant[1],BI_consv_hotspot=BI_consv_hotspot[1]), by=.(library, structural_alignment_pos, Pos)]

for(lib in libraries_binding_names){
  print(lib)
  print(paste0(BI_hotspots_dt_[library==lib & significant==T & BI_consv_hotspot==T]$Pos, collapse=","))
  print(paste0(BI_hotspots_dt_[library==lib & significant==T & BI_consv_hotspot==F]$Pos, collapse=","))
  print(paste0(BI_hotspots_dt_[library==lib & significant==F]$Pos, collapse=","))
}
```


# binding interface heatmaps 
```{r, fig.width=15, fig.height=7}
library(ggnewscale)
get_contacts_dt<-as.data.table(read.csv("/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis/01_structures/3DstructureMetrics/04-contacts/contacts_table_complete.csv", sep=","))

all_ddg_table[,mutation:=substr(id, nchar(id), nchar(id))]

get_contacts_dt_tmp<-get_contacts_dt[chain_1=="A" & chain_2=="B"]
setnames(get_contacts_dt, "Pos_1", "Pos")

get_contacts_dt$library<-lib_code_to_name(get_contacts_dt$library, libraries, libraries_names)

get_contacts_dt_<-left_join(get_contacts_dt, pdb_metrics_dt[,c("structural_alignment_pos", "Pos", "library")], by=c("Pos", "library"))
all_ddg_table$lib_name<-lib_code_to_name(all_ddg_table$library, libraries, libraries_names)
all_ddg_table$lib_name<-gsub("_", "-", gsub(" binding ", " | ", all_ddg_table$lib_name))
all_median_ddg_table$lib_name<-lib_code_to_name(all_median_ddg_table$library, libraries, libraries_names)
all_median_ddg_table$lib_name<-gsub("_", "-", gsub(" binding ", " | ", all_median_ddg_table$lib_name))
get_contacts_dt_$lib_name<-lib_code_to_name(get_contacts_dt_$library, libraries, libraries_names)
get_contacts_dt_$lib_name<-gsub("_", "-", gsub(" binding ", " | ", get_contacts_dt_$lib_name))


all_hotspots<-unique(BI_hotspots_dt[significant==T]$structural_alignment_pos)
p<-ggplot(all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,],aes(x=lib_name, y=mutation))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=ddg))+ 
  facet_wrap(~structural_alignment_pos, nrow=1)+
  theme_clean(base_size = 13)+
    scale_fill_gradient2(low="blue",mid="grey90",high="red",midpoint=0, na.value = "grey50")+
    scale_y_discrete(limits=c("distance",aa_order2))+
  scale_x_discrete(limits=unique(all_ddg_table[assay=="binding"]$lib_name))+
    theme( plot.title = element_text(size=35))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), 
           axis.ticks.x = element_blank(), 
           #panel.grid.minor = element_blank(), 
           axis.text.y = element_text(size=10, color=c("#FF7013",colors_aa), face="bold"), 
           plot.margin = unit(c(1, 0,0,0), "cm"),
           legend.text = element_text(size=15), 
           legend.title =  element_text(size=20)) + 
  geom_text(data = all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], size=5, aes(x=lib_name, y=WT_aa, label = WT_aa, color=WT_aa), show.legend=F)+
scale_color_manual(limits=aa_order2, values=colors_aa)+
  
  new_scale("fill")+
  geom_point(data = all_median_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=1, label = WT_aa, fill=scHAmin_ligand), shape=21, size=3)+
  scale_fill_gradient2("distance", low="#FF7013",mid="white",high="#36C6CC",midpoint=5, na.value = "grey50")+
  theme(legend.position = "right")+
  
  new_scale("fill")+
  geom_point(data=get_contacts_dt_[chain_1=="A" & chain_2=="B" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=residue_2_letter, fill=interaction_types), shape=23, size=2.5, alpha=1)

ggsave(paste0("Figs/Fig3/FigS3CBI_hotspots_heatmaps_legend.png"), width=18, height=7)

p+theme(legend.position = "none", strip.text = element_text(size=14, face="bold"), axis.text.y = element_text(size=14, face="bold"))
```



```{r, fig.width=20, fig.height=7}
p<-ggplot(all_ddg_table[assay=="binding" & structural_alignment_pos %in% BI_hotspots_dt_pdz_nohotspots,],aes(x=lib_name, y=Mut))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=ddg))+ 
  facet_wrap(~structural_alignment_pos, nrow=1)+
  theme_clean(base_size = 13)+
    scale_fill_gradient2(low="blue",mid="grey90",high="red",midpoint=0, na.value = "grey50")+
    scale_y_discrete(limits=c("distance",aa_order2))+
  scale_x_discrete(limits=unique(all_ddg_table[assay=="binding"]$lib_name))+
    theme( plot.title = element_text(size=35))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), 
           axis.ticks.x = element_blank(), 
           #panel.grid.minor = element_blank(), 
           axis.text.y = element_text(size=10, color=c("#FF7013",colors_aa), face="bold"), 
           plot.margin = unit(c(1, 0,0,0), "cm"),
           legend.text = element_text(size=15), 
           legend.title =  element_text(size=20)) + 
  geom_text(data = all_ddg_table[assay=="binding" & structural_alignment_pos %in% BI_hotspots_dt_pdz_nohotspots,], size=5, aes(x=lib_name, y=WT_aa, label = WT_aa, color=WT_aa), show.legend=F)+
scale_color_manual(limits=aa_order2, values=colors_aa)+
  
  new_scale("fill")+
  geom_point(data = all_median_ddg_table[assay=="binding" & structural_alignment_pos %in% BI_hotspots_dt_pdz_nohotspots,], aes(x=lib_name, y=1, label = WT_aa, fill=scHAmin_ligand), shape=21, size=3)+
  scale_fill_gradient2("distance", low="#FF7013",mid="white",high="#36C6CC",midpoint=5, na.value = "grey50")+
  theme(legend.position = "right")+
  
  new_scale("fill")+
  geom_point(data=get_contacts_dt_[chain_1=="A" & chain_2=="B" & structural_alignment_pos %in% BI_hotspots_dt_pdz_nohotspots,], aes(x=lib_name, y=residue_2_letter, fill=interaction_types), shape=23, size=2.5, alpha=1)

p+theme(legend.position = "none", strip.text = element_text(size=14, face="bold"), axis.text.y = element_text(size=14, face="bold"))
ggsave(paste0("Figs/Fig3/FigS3CBI_nonhotspots_heatmaps.png"), width=18, height=7)
```


# enrichment of non-hotspots in b2-b3 loop
```{r}
all_non_hotspots<-BI_hotspots_dt_pdz[num_pdzs==0]$structural_alignment_pos
all_non_hotspots_dt<-data.table(structural_alignment_pos=all_bi)
all_non_hotspots_dt[, b2_b3_loop:=structural_alignment_pos %in% c(30:44)]
all_non_hotspots_dt[, no_htspot:=structural_alignment_pos %in% all_non_hotspots]

fisher.test(table(all_non_hotspots_dt$b2_b3_loop, all_non_hotspots_dt$no_htspot), alternative = "greater")
```

# save the hotspots definition table supplementary 6
```{r}
BI_hotspots_dt_complete<-left_join(BI_hotspots_dt[,.(library, lib_name, structural_alignment_pos, ddg, std_ddg, WT_aa, rsasa, binding_interface_contacts, class_BI, hotspot_class, Uniprot_pos_ref.x)], test_results)


BI_hotspots_dt_complete<-BI_hotspots_dt_complete[, .(mean_ddg=mean(ddg), rsasa=rsasa[1],p_value=p_value[1], p_adjusted=p_adjusted[1], median_ddg=median(ddg), BI_hotspot_class=hotspot_class[1]), by=.(lib_name, structural_alignment_pos, Uniprot_pos_ref.x)][,class:="binding_interface"]

BI_hotspots_dt_complete$lib_name<-tolower(gsub("-", "_",gsub("\n", "_", BI_hotspots_dt_complete$lib_name)))
colnames(BI_hotspots_dt_complete)[3]<-"Uniprot_pos_ref"

fwrite(BI_hotspots_dt_complete, paste0("supplementary_tables/Supplementary_table6_BI.txt"))
```









