---
title: "PDZ Binding interfaces"
author: "Aina Martí-Aranda"
date: "June 2025"
output:
  html_document:
    toc: true
    theme: cerulean
    toc_float: true
    toc_collapsed: true
---


# initializing
```{r setup}
base_dir<-"/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis_github/"
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

# load some variables
load('source_variables.RData')
```

## libraries used for the analysis
```{r message=FALSE, warning=FALSE}
library(car)
library(systemfonts)
library(plyr)
library(ggiraph)
library(ggiraphExtra)
library(rstatix)
library(data.table)
library(dplyr)
library(seqinr)
library(stringr)
library(ggplot2)
library(Matrix)
library(viridis)
library(ggpubr)
library(readxl)
library(stringi)
library(GGally)
library(readxl)
library(modelr)
library(cowplot)
library(Hmisc)
library(ggrepel)
library(dplyr) 
library(ddpcr)
library(ggpointdensity)
library(tidyr)
library(magick)
library(ggpointdensity)
library(ggpmisc)
library(ggridges)
```


# load some tables
```{r}
all_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_ddg_table.csv")))
all_median_ddg_table<-as.data.table(read.csv(paste0("tmp_tables/all_median_ddg_table.csv")))
pdb_metrics_dt<-as.data.table(read.csv("data/annotations/structure_metrics.csv", sep=" "))
aln_table<-as.data.table(read.table(paste0("data/annotations/structural_alignment_tcoffee.csv")))
get_contacts_dt<-as.data.table(read.csv("data/annotations/tmp_files/contacts_table_complete.csv", sep=","))
```


# preparing some variables
```{r}
#preparing some data
all_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

all_median_ddg_table[is.na(binding_interface_contacts), binding_interface_contacts:=F]
all_median_ddg_table[is.na(core_bi_contacts), core_bi_contacts:=F]

#prepare binding tables
all_ddg_table_binding<-all_ddg_table[assay=="binding"]
all_median_ddg_table_binding<-all_median_ddg_table[assay=="binding"]

#list all + common binding interface positions
all_bi<-unique(all_ddg_table_binding[ binding_interface_contacts==T]$structural_alignment_pos)
core_bi<-unique(all_ddg_table_binding[core_bi_contacts==T]$structural_alignment_pos)

# saving this lists
save(all_bi, file = paste0("tmp_tables/all_bi.Rdata"))
save(core_bi, file = paste0("tmp_tables/core_bi.Rdata"))

#classify these positions
all_ddg_table_binding[,class_BI:="outside_BI"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi,class_BI:="some_BI, contact"]
all_ddg_table_binding[structural_alignment_pos %in% all_bi & binding_interface_contacts==F,class_BI:="some_BI, NOcontact"]
all_ddg_table_binding[structural_alignment_pos %in% core_bi,class_BI:="core_BI"]
```


# Fig 3.a: BI distribution of effects

```{r}
all_ddg_table_binding$lib_name<-lib_code_to_name(all_ddg_table_binding$library, libraries_binding_names, libraries_names_plots[which(libraries %in% libraries_binding)])
all_ddg_table_binding$lib_name<-gsub(" binding ", "\n", all_ddg_table_binding$lib_name)
all_ddg_table_binding$lib_name<-gsub("_", "-", all_ddg_table_binding$lib_name)

p<-ggplot(all_ddg_table_binding[structural_alignment_pos %in% all_bi], aes(x=ddg))+
  #geom_density(scale="width", size=1, aes(color=library))+#color="grey30", fill="grey80",
  theme_classic()+
  xlab("∆∆Gb in binding\ninterface positions")+
  theme(#legend.key.size = unit(0.3, 'cm'),
        #legend.text = element_text(size=7),
        legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x=element_text(angle=90, vjust=0.5, hjust = 1, size=7))+
  xlim(c(-1.5,3))+
  #scale_color_brewer(palette="Set2")+
  #scale_fill_brewer(palette="Set2")+
  geom_density_ridges2(aes(y=lib_name), scale = 1.3, alpha=0.8, fill="grey80", color="grey50")+coord_flip()

p+theme(axis.text.x=element_text(size=9,face="bold", color="black"), 
        axis.text.y=element_text(size=13,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))
ggsave(paste0("Figs/Fig3/Fig3AddGb_distribution_BI.png"), width=6, height=3)
```

## Identify hotspots
Defining hotspots as residues where mutations cause changes in binding energy larger than the median of all interface residues (0.75 kcal/mol)
```{r, fig.width=10, fig.height=10}
# all candidate positions
plot_dt<-all_ddg_table_binding[binding_interface_contacts==T]
summary(plot_dt$ddg,na.rm = T)

# Calculate the median of all BI residues
median_BI <- median(all_ddg_table_binding[binding_interface_contacts == T]$ddg, na.rm = TRUE)

# Filter groups with at least 2 observations
filtered_data <- plot_dt[, 
  .N, by = .(library, structural_alignment_pos)][N >= 2]

# Merge the filtered groups back with the original data
valid_data <- plot_dt[library %in% filtered_data$library &  structural_alignment_pos %in% filtered_data$structural_alignment_pos,]


# Perform the t-test only for valid groups
test_results <- valid_data[
  !is.na(ddg),  # Ensure ddg is not NA
  .(
    p_value = if (.N > 1 && length(unique(ddg)) > 1) {
      t.test(ddg, mu = median_BI, alternative = "greater")$p.value
    } else {
      NA_real_  # Use numeric NA for consistency
    }
  ),
  by = .(library, structural_alignment_pos)
]

test_results[, p_adjusted := p.adjust(p_value, method = "fdr")]
significant_distributions<-test_results[p_adjusted<0.1] # save the significant ones in a new distribution
significant_distributions[,significant:=T]
plot_dt<-left_join(plot_dt, significant_distributions)

# CONSERVED binding interface hotspots
# define the conserved binding interface hotspots those with >=4 libraries significantly above the median
BI_hotspots<-significant_distributions[significant==T, .N, by=c("structural_alignment_pos")][N>=4]$structural_alignment_pos
save(BI_hotspots, file = paste0("tmp_tables/BI_hotspots.Rdata"))

# Add a new column to label significant facets
all_ddg_table_binding[, highlight := ifelse(structural_alignment_pos %in% BI_hotspots, "highlight", "normal")]

plot_dt[is.na(significant), significant:=F]
plot_dt[, BI_consv_hotspot:=(structural_alignment_pos %in% BI_hotspots)]
BI_hotspots_dt<-plot_dt
fwrite(BI_hotspots_dt, paste0("tmp_tables/BI_hotspots_dt.csv"))

# prepare library names annotations
plot_dt$lib_name<-lib_code_to_name(plot_dt$library, libraries_binding_names, libraries_names_plots[which(libraries %in% libraries_binding)])
plot_dt$lib_name<-gsub(" binding ", " | ", plot_dt$lib_name)
plot_dt$lib_name<-gsub("_", "-", plot_dt$lib_name)

# annotate secondary structure labels
plot_dt[, region:=""]
plot_dt[structural_alignment_pos %in% seq(24,30), region:="β2"]
plot_dt[structural_alignment_pos %in% c(24,25,26), region:="carboxylate binding loop"]
plot_dt[structural_alignment_pos %in% seq(30,45), region:="β2-β3 loop"]
plot_dt[structural_alignment_pos %in% seq(46,52), region:="β3"]
plot_dt[structural_alignment_pos %in% seq(82,91), region:="α3"]
plot_dt[, region:=structural_alignment_pos]



p<-ggplot(plot_dt)+
  geom_hline(aes(yintercept = median(plot_dt$ddg,na.rm = T)), color="red3")+
  geom_hline(aes(yintercept = 0), color="grey", linetype="dashed")+
  geom_boxplot(aes(x=factor(structural_alignment_pos), y=ddg, fill=lib_name, color=significant))+
  scale_fill_brewer(palette="Set2")+
  theme_pubr()+
  scale_color_manual("Fdr<0.1",values=c("grey80", "black"))+
  #facet_grid(rows = vars(region), cols = vars(factor(structural_alignment_pos)), scales = "free_x")+
  facet_wrap(~region+factor(structural_alignment_pos), scales="free_x", ncol=7)+#, ncol=13)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x =element_blank(),
        legend.position="right",
    strip.text = element_text(
      color="black",
      size = 10,
      face = "bold"
    )
  ) +ylab("∆∆G")+xlab("")#+
  #geom_point(data=plot_dt[BI_consv_hotspot==T,], aes(x=1, y=4), shape=23, size=3.5, fill="#FF7013")+xlab("")+ylab("∆∆Gb")

fwrite(plot_dt, paste0("tmp_tables/BI_hotspots_dt.csv"))

p+theme(legend.position="none")
ggsave(paste0("Figs/Fig3/Fig3Bdefine_hotspots.png"), width=10, height=10)


```


# percentage of hotspots per interaction
A median of six hotspots per interface (false discovery rate (FDR, Benjamini-Hochberg method) < 0.1, one-sided t-test). This is a median of 41.7% of interface residues (range 29.4% to 61.5% Fig. 3b).
```{r}
BI_hotspots_dt_<-BI_hotspots_dt[, .(significant=significant[1]), by=.(structural_alignment_pos, library)]
BI_hotspots_dt_[,pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_counts_<-BI_hotspots_dt_[, .(.N, N_sign=sum(significant)), by=.(library)]
BI_hotspots_dt_counts_[,percentage:=N_sign/N*100]
summary(BI_hotspots_dt_counts_$percentage)
```



Only two residues are hotspots in all seven interactions (structural alignment positions 25 in the carboxylate binding loop, and 86 in the α2 helix), and an additional four are conserved in more than half of the interactions (positions 24 in the carboxylate binding loop, 26 and 28 in β2, and 82 in α2, Fig. 3b,c,d).
```{r}
BI_hotspots_dt_<-BI_hotspots_dt[, .(significant=significant[1]), by=.(structural_alignment_pos, library)]
BI_hotspots_dt_[,pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_libraries<-BI_hotspots_dt_[, .(N=.N, num_hotspots=sum(significant)), by=.(structural_alignment_pos)]
BI_hotspots_dt_pdz<-BI_hotspots_dt_[significant==T, .(num_pdzs=length(unique(pdz))), by=.(structural_alignment_pos)]
BI_hotspots_dt_libraries[num_hotspots==7] # which are hotspots in all interactions
BI_hotspots_dt_pdz[num_pdzs==5] # which are hotspots in all 5 PDZs
BI_hotspots_dt_pdz[num_pdzs>=3] # which are hotspots in half of the PDZs
```
A total of 15 interface residues are never classified as hotspots, including eight (positions 23, 29, 32, 34, 41, 50, 83, and 90) that make contacts with a ligand in multiple PDZ domains (Fig. 3b, extended Data Fig. 3a,c). 
```{r}
#never classified as hotspots
BI_hotspots_dt_[, .(num_pdzs=sum(significant)), by=.(structural_alignment_pos)][num_pdzs==0]

# never classified as hotspots when interacting in several PDZs
BI_hotspots_dt_[, pdz:=sub("^(([^_]+)_([^_]+))_.*", "\\1", library)]
BI_hotspots_dt_[, .(num_pdzs=sum(significant), total_contacts_pdzs=length(unique(pdz))), by=.(structural_alignment_pos)][num_pdzs==0 & total_contacts_pdzs>1][order(structural_alignment_pos)]
```



plot all the counts
```{r warning=FALSE, fig.width=3, fig.height=3}
# now I add the ones that are 0
BI_hotspots_dt_none_sign<-BI_hotspots_dt_[, .(num_pdzs=sum(significant)), by=.(structural_alignment_pos)][num_pdzs==0]
BI_hotspots_dt_pdz<-rbind(BI_hotspots_dt_pdz, BI_hotspots_dt_none_sign)
BI_hotspots_dt_pdz<-rbind(BI_hotspots_dt_pdz, data.table(structural_alignment_pos=0, num_pdzs=2))

ggplot(BI_hotspots_dt_pdz)+
  geom_bar(aes(x=factor(num_pdzs, levels=c(0,1,2,3,4,5)), fill=factor(num_pdzs)), stat="count")+
  #geom_bar(aes(x=factor(N_total), y=V1), stat="identity", bins=7, fill="black")+
  scale_fill_manual(values=c("grey", "black","white", "brown1","brown1", "brown1", "brown1", "brown1"))+
   ylab("Position count")+
  #scale_fill_manual(values=c(rep("grey", 3), "red"))+
  xlab("# PDZs with hotspot")+
  theme_classic()+theme(legend.position="none")+
  #scale_x_discrete(breaks=-1:5, limits=-1:5)+
  theme(axis.text.x=element_text(size=10,face="bold", color="black"), 
        axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"),
        axis.title.x=element_text(size=10,face="bold", color="black"))
ggsave(paste0("Figs/Fig3/Fig3Clibraries_per_hotspot.png"), width=2.5, height=2.5)
```


# Position contacts of hotspots (histogram)
Hotspots are enriched in contacts with ligand positions 0 or -2, the two aa positions used to define the three main classes of PDZ domain ligands (OR=13.71, P=2.61x10-7, one-sided Fisher’s Exact Test (FET) Fig. 3e)
```{r}
get_contacts_dt_aln_ligand<-get_contacts_dt[,Pos:=Pos_1][chain_2=="B" & chain_1=="A", .(lig_contacts=paste(Pos_2-9, collapse=",")), by=.(Pos, library)]

BI_hotspots_dt_w_contacts<-left_join(get_contacts_dt_aln_ligand, BI_hotspots_dt[,.(significant=significant[1]) ,by=.(library, Pos, structural_alignment_pos)][,.(library, Pos, structural_alignment_pos, significant)])

# all of the conserved residues conract position 0 and/or -2
BI_hotspots_dt_w_contacts[significant==T & structural_alignment_pos %in% BI_hotspots]

BI_hotspots_dt_w_contacts[, has_0_or_neg2 := grepl("(^|,)(-2|0)(,|$)", lig_contacts)]
BI_hotspots_dt_w_contacts[, conserved_hotspot:=structural_alignment_pos %in% BI_hotspots]

#enrichment of has_0_or_neg2 in overall hotspots
fisher.test(table(BI_hotspots_dt_w_contacts$has_0_or_neg2, BI_hotspots_dt_w_contacts$significant), alternative = "greater" )
```
```{r}
# plot the number of contacts
BI_hotspots_dt_w_contacts[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt_w_contacts[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt_w_contacts[conserved_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]

dt_expanded <- BI_hotspots_dt_w_contacts[, .(lig_pos = unlist(strsplit(lig_contacts, ","))), by = .(hotspot_class)]
ligand_contact_counts <- dt_expanded[, .N, by = .(hotspot_class, lig_pos)][order(hotspot_class, lig_pos)]

ggplot(ligand_contact_counts, aes(x=factor(as.numeric(lig_pos)), y=N, fill=hotspot_class))+
  geom_bar(stat="identity")+
  facet_wrap(~factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), ncol=1)+
  scale_x_discrete(limits = factor(c(0:-8)))+xlab("ligand position")+ylab("number of contacts")+
  theme_classic()+
  theme(legend.position = "none", strip.text = element_blank())+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  theme(axis.text.x=element_text(size=10,face="bold", color="black"), 
        axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"),
        axis.title.x=element_text(size=10,face="bold", color="black"))
ggsave(paste0("Figs/Fig3/Fig3Ehotspot_contacts.png"), width=3, height=3)
```


# conserved hotapots are highly conserved structural contacts
The conserved hotspot positions also have higher sequence and structural conservation (Fig. 3f,g), and have highly conserved structural contacts, with five out of six positions contacting the ligand in all seven interactions (OR=60.54, P=5.3x10-04, one-sided FET).
```{r}
#enrichment in core_bi residues
dt_tmp<-data.table(all_bi=all_bi)
dt_tmp[all_bi %in% BI_hotspots, BI_hotspot:=T]
dt_tmp[!all_bi %in% BI_hotspots, BI_hotspot:=F]
dt_tmp[all_bi %in% core_bi, core_BI:=T]
dt_tmp[!all_bi %in% core_bi, core_BI:=F]
fisher.test(table(dt_tmp$core_BI, dt_tmp$BI_hotspot), alternative = "greater")
```


# conservarion in sequence
The conserved hotspot positions also have higher sequence and structural conservation (Fig. 3f,g), and have highly conserved structural contacts, with five out of six positions contacting the ligand in all seven interactions (OR=60.54, P=5.3x10-04, one-sided FET).
```{r, fig.width=4, fig.height=4}
BI_hotspots_dt_tmp<-BI_hotspots_dt
BI_hotspots_dt_tmp[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt_tmp[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt_tmp[BI_consv_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple<-BI_hotspots_dt_tmp[, .(significant=significant[1], consv=consv[1], hotspot_class=hotspot_class[1]), by=.(library, structural_alignment_pos)]

BI_hotspots_dt_simple<-BI_hotspots_dt_simple[,.(hotspot_class=hotspot_class[1], consv=consv[1]), by=.(structural_alignment_pos)]
BI_hotspots_dt_simple[structural_alignment_pos %in% BI_hotspots_dt_simple, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple[structural_alignment_pos %in% BI_hotspots_dt_none_sign$structural_alignment_pos, hotspot_class:="non-Hotspot"]


ggplot(BI_hotspots_dt_simple, aes(x=factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), y=consv, fill=hotspot_class, color=hotspot_class))+
  #geom_hline(aes(yintercept = 1), color="grey")+
  #geom_hline(aes(yintercept = 0), color="grey")+
  geom_violin(scale="width", alpha=0.7)+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  scale_color_manual(values=c("black", "brown1", "grey"))+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  theme_classic()+
  ylab("Sequence conservation score")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), axis.title.x=element_blank(), legend.position="none")+
  stat_compare_means(comparisons = list(c("Hotspot in >2/5 PDZs", "Hotspot"), c("Hotspot", "non-Hotspot")), aes(label = after_stat(p.signif)))+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))

ggsave(paste0("Figs/Fig3/Fig3Fseq_consv_hotspots.png"), width=2.3, height=2.3)
```



# conservation in structure
The conserved hotspot positions also have higher sequence and structural conservation (Fig. 3f,g), and have highly conserved structural contacts, with five out of six positions contacting the ligand in all seven interactions (OR=60.54, P=5.3x10-04, one-sided FET).  

Here I have superimposed the structures using Matchmaker from ChimeraX,  
And then manually computed the RMSD of the residue subsets of interest
```{r warning=FALSE, fig.width=4, fig.height=4}
#load models
library(bio3d)
model1 <- read.pdb(paste0("data/structures/superimposed/superimposed_dlg4_pdz3_cript.pdb.pdb"))
model2 <- read.pdb(paste0("data/structures/superimposed/superimposed_dlg4_pdz2_cript.pdb.pdb"))
model3 <- read.pdb(paste0( "data/structures/superimposed/superimposed_dlg4_pdz2_nmdar2a.pdb.pdb"))
model4 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz2_cep164.pdb.pdb"))
model5 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz1_dlgap.pdb.pdb"))
model6 <- read.pdb(paste0("data/structures/superimposed/superimposed_nhrf3_pdz1_uhmk1.pdb.pdb"))
model7 <- read.pdb(paste0("data/structures/superimposed/superimposed_erbin_pdz1_cript.pdb.pdb"))
models<-list(model1, model2, model3, model4, model5, model6, model7)


#define the function that will compute the local rmsd (actually local euclidean distances)
calculate_local_rmsd <- function(model1, model2, pair) {
  index1<-model1$atom$resno==pair[1] &  model1$atom$elety=="CA" & model1$atom$chain=="A"
  index2<-model2$atom$resno==pair[2] &  model2$atom$elety=="CA" & model2$atom$chain=="A"
  coord1 <- c(model1$atom$x[index1], model1$atom$y[index1], model1$atom$z[index1])
  coord2 <- c(model2$atom$x[index2], model2$atom$y[index2], model2$atom$z[index2])
   # Euclidean distance (squared differences summed)
  rmsd <- sqrt(sum((coord1 - coord2)^2))   # Root Mean Square Deviation
  return(rmsd)
}

# build the local (euclidean) distances between aligned residue's CA atoms
combinations<-t(combn(1:7, 2))
rmsd_comparisons<-data.table()
for(i in 1:nrow(combinations)){
  pair<-combinations[i,]
  libraries_<-libraries_binding_names[pair]
  model1<-models[pair[1]]
  model2<-models[pair[2]]
  
  #paired positions
  paired_dt<-t(reshape(all_median_ddg_table_binding[library %in% libraries_, c("structural_alignment_pos", "Pos", "library")], timevar="structural_alignment_pos", idvar="library", direction="wide"))
  aln_positions<-as.numeric(lapply(strsplit(rownames(paired_dt[-1,]), "[.]"), "[[", 2))
  paired_dt<-data.table(paired_dt[-1,])
  paired_dt$structural_alignment_pos<-aln_positions
  paired_dt <- na.omit(paired_dt)
  
  for(b_a_p in paired_dt$structural_alignment_pos){
    positions<-as.numeric(c(paired_dt[structural_alignment_pos==b_a_p]$V1,paired_dt[structural_alignment_pos==b_a_p]$V2))
    local_rmsd <- calculate_local_rmsd(model1[[1]], model2[[1]], positions)
    rmsd_comparisons<-rbind(rmsd_comparisons, data.table( paste0(libraries_, collapse="."), libraries_[1], libraries_[2], local_rmsd, b_a_p))
  }

}

colnames(rmsd_comparisons)<-c("lib_pair", "lib1", "lib2", "local_rmsd", "structural_alignment_pos")
rmsd_comparisons<-left_join(rmsd_comparisons[,library:=lib1], BI_hotspots_dt[,.(library, structural_alignment_pos, significant)])[, hotspot1:=significant]
rmsd_comparisons<-left_join(rmsd_comparisons[,library:=lib2], BI_hotspots_dt[,.(library, structural_alignment_pos, significant)])[, hotspot2:=significant]

rmsd_comparisons_studyBI<-rmsd_comparisons
rmsd_comparisons_studyBI[is.na(hotspot1) & is.na(hotspot2), hotspot:="non-hotspot"]
rmsd_comparisons_studyBI[!is.na(hotspot1) | !is.na(hotspot2), hotspot:="hotspot"]
rmsd_comparisons_studyBI[!is.na(hotspot1) & !is.na(hotspot2), hotspot:="both hotspot"]
rmsd_comparisons_studyBI[structural_alignment_pos %in% BI_hotspots & hotspot!="non-hotspot" & hotspot!="hotspot", hotspot:="conserved hotspot"]


rmsd_results <- rmsd_comparisons_studyBI[, .(
  non_hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "non-hotspot"])^2)) else NA_real_,
  consv_hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "conserved hotspot"])^2)) else NA_real_,
  hotspot = if (.N > 0) sqrt(mean((local_rmsd[hotspot == "both hotspot"])^2)) else NA_real_
), by=lib_pair]

# annotate libraries
dt_tmp<-melt(rmsd_results, id.var="lib_pair")
dt_tmp[, class2 := as.integer(grepl("4", lib_pair))]

dt_tmp[, lib1 := sub("\\..*", "", lib_pair)]      # before the first dot
dt_tmp[, lib2 := sub("^[^.]*\\.", "", lib_pair)]  # after the first dot

dt_tmp[sub("^(([^_]+)_([^_]+))_.*", "\\1", lib1)!=sub("^(([^_]+)_([^_]+))_.*", "\\1", lib2), diff_pdz:="different PDZ"] #keep only comparisons between different PDZs
dt_tmp[sub("^(([^_]+)_([^_]+))_.*", "\\1", lib1)==sub("^(([^_]+)_([^_]+))_.*", "\\1", lib2), diff_pdz:="same PDZ"]


# plot the distributions per group
ggplot(dt_tmp[diff_pdz=="different PDZ"], aes(x=factor(variable, levels=c("consv_hotspot","hotspot", "non_hotspot"), labels=c("BI conserved\nhotspot","hotspot", "non-hotspot")), y=value, fill=variable, color=variable))+
  geom_violin(scale="width", alpha=0.7)+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  #geom_point(color="black", alpha=0.8)+
  stat_compare_means(comparisons=list(c("BI conserved\nhotspot","hotspot"), c("non-hotspot","hotspot")), size=5, aes(label=after_stat(p.signif)), paired=T)+
  scale_fill_manual(values=c("grey","brown1", "black"))+
  scale_color_manual(values=c("grey","brown1", "black"))+
  theme_classic()+
  #geom_line(aes(group=V1), color="grey30", size=0.1)+
  ylim(c(0,6.8))+
  #facet_grid(~diff_pdz)+
  ylab("RMSD")+#ggtitle("strutctural conservation of residues in the BI")+
  theme(legend.position = "none", axis.title.x=element_blank())+
  theme(legend.position="none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))

ggsave(paste0("Figs/Fig3/Fig3Gstructural_conservation_hotspot.png"), width=2.5, height=2.5)

```

# rsasa of the hotspots

The conserved hotspots are also buried deeper in the hydrophobic binding pocket (median rSASA = 0.14 vs. 0.33 for conserved and non-conserved hotspots, respectively, Wilcoxon rank-sum test P=5.8x10-03; median rSASA for non-hotspots=0.58, P=1.5x10-14, Fig. 3h)

```{r, fig.width=4, fig.height=4}
BI_hotspots_dt[, hotspot_class:="non-Hotspot"]
BI_hotspots_dt[significant==T, hotspot_class:="Hotspot"]
BI_hotspots_dt[BI_consv_hotspot==T, hotspot_class:="Hotspot in >2/5 PDZs"]
BI_hotspots_dt_simple<-BI_hotspots_dt[, .(significant=significant[1], rsasa=rsasa[1], hotspot_class=hotspot_class[1]), by=.(library, structural_alignment_pos)]

ggplot(BI_hotspots_dt_simple, aes(x=factor(hotspot_class, levels=c("Hotspot in >2/5 PDZs", "Hotspot", "non-Hotspot")), y=rsasa, fill=hotspot_class, color=hotspot_class))+
  geom_hline(aes(yintercept = 0.25), color="grey")+
  #geom_hline(aes(yintercept = 0), color="grey")+
  geom_violin(scale="width", alpha=0.7)+
  scale_fill_manual(values=c("black", "brown1", "grey"))+
  scale_color_manual(values=c("black", "brown1", "grey"))+
  geom_point(position="jitter")+
  geom_boxplot(width=0.2, fill="white", color="black", outliers=F)+
  theme_classic()+
  ylab("RSASA")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), axis.title.x=element_blank(), legend.position="none")+
  stat_compare_means(comparisons = list(c("Hotspot in >2/5 PDZs", "Hotspot"), c("Hotspot in >2/5 PDZs", "non-Hotspot"), c("Hotspot", "non-Hotspot")), aes(label = after_stat(p.signif)))+ylim(c(0,1.4))+
  theme(axis.text.y=element_text(size=10,face="bold", color="black"),
        axis.title.y=element_text(size=10,face="bold", color="black"))

ggsave(paste0("Figs/Fig3/Fig3Hrsasa_hotspots.png"), width=2.5, height=2.5)

summary(BI_hotspots_dt_simple[hotspot_class=="Hotspot in >2/5 PDZs"]$rsasa)
summary(BI_hotspots_dt_simple[hotspot_class=="Hotspot"]$rsasa)
summary(BI_hotspots_dt_simple[hotspot_class=="non-Hotspot"]$rsasa)
```





# chimera hotspots figures
This loop prints the raw indexes of each interactions that are hotspots.  
The first one is the conserved hotspots.  
Second string are the non conserved hotspots. 
Finally, the BI residues non-hotspots.  
Used to select the residues in chimera to highlight them
```{r}
BI_hotspots_dt_<-BI_hotspots_dt[,.(significant=significant[1],BI_consv_hotspot=BI_consv_hotspot[1]), by=.(library, structural_alignment_pos, Pos)]

for(lib in libraries_binding_names){
  print(lib)
  print(paste0(BI_hotspots_dt_[library==lib & significant==T & BI_consv_hotspot==T]$Pos, collapse=","))
  print(paste0(BI_hotspots_dt_[library==lib & significant==T & BI_consv_hotspot==F]$Pos, collapse=","))
  print(paste0(BI_hotspots_dt_[library==lib & significant==F]$Pos, collapse=","))
}
```



# binding interface heatmaps 
```{r, fig.width=15, fig.height=7}
library(ggnewscale)
get_contacts_dt<-as.data.table(read.csv("/Users/aa33/Library/CloudStorage/OneDrive-WellcomeSangerInstitute/001-Projects/03-PDZ_Homologs/02-DryLab/001-data_analysis/data_analysis/01_structures/3DstructureMetrics/04-contacts/contacts_table_complete.csv", sep=","))

all_ddg_table[,mutation:=substr(id, nchar(id), nchar(id))]

get_contacts_dt_tmp<-get_contacts_dt[chain_1=="A" & chain_2=="B"]
setnames(get_contacts_dt, "Pos_1", "Pos")

get_contacts_dt_<-left_join(get_contacts_dt, pdb_metrics_dt[,c("structural_alignment_pos", "Pos", "library")], by=c("Pos", "library"))
all_ddg_table$lib_name<-lib_code_to_name(all_ddg_table$library, libraries, libraries_names)
all_ddg_table$lib_name<-gsub("_", "-", gsub(" binding ", " | ", all_ddg_table$lib_name))
all_median_ddg_table$lib_name<-lib_code_to_name(all_median_ddg_table$library, libraries, libraries_names)
all_median_ddg_table$lib_name<-gsub("_", "-", gsub(" binding ", " | ", all_median_ddg_table$lib_name))
get_contacts_dt_$lib_name<-lib_code_to_name(get_contacts_dt_$library, libraries, libraries_names)
get_contacts_dt_$lib_name<-gsub("_", "-", gsub(" binding ", " | ", get_contacts_dt_$lib_name))


all_hotspots<-unique(BI_hotspots_dt[significant==T]$structural_alignment_pos)
p<-ggplot(all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,],aes(x=lib_name, y=mutation))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=ddg))+ 
  facet_wrap(~structural_alignment_pos, nrow=1)+
  theme_clean(base_size = 13)+
    scale_fill_gradient2(low="blue",mid="grey90",high="red",midpoint=0, na.value = "grey50")+
    scale_y_discrete(limits=c("distance",aa_order2))+
  scale_x_discrete(limits=unique(all_ddg_table[assay=="binding"]$lib_name))+
    theme( plot.title = element_text(size=35))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), 
           axis.ticks.x = element_blank(), 
           #panel.grid.minor = element_blank(), 
           axis.text.y = element_text(size=10, color=c("#FF7013",colors_aa), face="bold"), 
           plot.margin = unit(c(1, 0,0,0), "cm"),
           legend.text = element_text(size=15), 
           legend.title =  element_text(size=20)) + 
  #ggtitle("CRIPT binding libraries")+
    geom_text(data = all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], size=5, aes(x=lib_name, y=WT_aa, label = WT_aa, color=WT_aa), show.legend=F)+
scale_color_manual(limits=aa_order2, values=colors_aa)+
  
  new_scale("fill")+
  geom_point(data = all_median_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=1, label = WT_aa, fill=scHAmin_ligand), shape=21, size=3)+
  scale_fill_gradient2("distance", low="#FF7013",mid="white",high="#36C6CC",midpoint=5, na.value = "grey50")+
  theme(legend.position = "right")+
  
  new_scale("fill")+
  geom_point(data=get_contacts_dt_[chain_1=="A" & chain_2=="B" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=residue_2_letter, fill=interaction_types), shape=23, size=2.5, alpha=1)

ggsave(paste0("Figs/Fig3/FigS3CBI_hotspots_heatmaps_legend.png"), width=18, height=7)

p+theme(legend.position = "none", strip.text = element_text(size=14, face="bold"), axis.text.y = element_text(size=14, face="bold"))
#ggsave(paste0(results_dir_manuscript,"/Fig3/BI_hotspots_heatmaps.png"), width=18, height=8)
```



```{r, fig.width=20, fig.height=7}
all_hotspots<-BI_hotspots_dt_pdz[num_pdzs==0]$structural_alignment_pos
p<-ggplot(all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,],aes(x=lib_name, y=mutation))+
    xlab("")+
    ylab("")+
    geom_tile(aes(fill=ddg))+ 
  facet_wrap(~structural_alignment_pos, nrow=1)+
  theme_clean(base_size = 13)+
    scale_fill_gradient2(low="blue",mid="grey90",high="red",midpoint=0, na.value = "grey50")+
    scale_y_discrete(limits=c("distance",aa_order2))+
  scale_x_discrete(limits=unique(all_ddg_table[assay=="binding"]$lib_name))+
    theme( plot.title = element_text(size=35))+
    theme( panel.spacing.x = unit(0.1, "lines"), 
           axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), 
           axis.ticks.x = element_blank(), 
           #panel.grid.minor = element_blank(), 
           axis.text.y = element_text(size=10, color=c("#FF7013",colors_aa), face="bold"), 
           plot.margin = unit(c(1, 0,0,0), "cm"),
           legend.text = element_text(size=15), 
           legend.title =  element_text(size=20)) + 
  #ggtitle("CRIPT binding libraries")+
    geom_text(data = all_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], size=5, aes(x=lib_name, y=WT_aa, label = WT_aa, color=WT_aa), show.legend=F)+
scale_color_manual(limits=aa_order2, values=colors_aa)+
  
  new_scale("fill")+
  geom_point(data = all_median_ddg_table[assay=="binding" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=1, label = WT_aa, fill=scHAmin_ligand), shape=21, size=3)+
  scale_fill_gradient2("distance", low="#FF7013",mid="white",high="#36C6CC",midpoint=5, na.value = "grey50")+
  theme(legend.position = "right")+
  
  new_scale("fill")+
  geom_point(data=get_contacts_dt_[chain_1=="A" & chain_2=="B" & structural_alignment_pos %in% all_hotspots,], aes(x=lib_name, y=residue_2_letter, fill=interaction_types), shape=23, size=2.5, alpha=1)

p+theme(legend.position = "none", strip.text = element_text(size=14, face="bold"), axis.text.y = element_text(size=14, face="bold"))
ggsave(paste0("Figs/Fig3/FigS3CBI_nonhotspots_heatmaps.png"), width=18, height=7)
```


# enrichment of non-hotspots in b2-b3 loop
These non-hotspot contacts are more solvent accessible (Fig. 3h) and they are particularly enriched in the highly variable β2–β3 loop31 (OR=6.23, P=0.04, one-sided FET, Fig. 3b and Extended Data Fig. 3a).
```{r}
all_non_hotspots<-BI_hotspots_dt_pdz[num_pdzs==0]$structural_alignment_pos
all_non_hotspots_dt<-data.table(structural_alignment_pos=all_bi)
all_non_hotspots_dt[, b2_b3_loop:=structural_alignment_pos %in% c(30:44)]
all_non_hotspots_dt[, no_htspot:=structural_alignment_pos %in% all_non_hotspots]

fisher.test(table(all_non_hotspots_dt$b2_b3_loop, all_non_hotspots_dt$no_htspot), alternative = "greater")
```

# save the hotspots definition table
```{r}
BI_hotspots_dt_complete<-left_join(BI_hotspots_dt[,.(library, lib_name, structural_alignment_pos, ddg, std_ddg, WT_aa, rsasa, binding_interface_contacts, class_BI, hotspot_class, Uniprot_pos_ref.x)], test_results)


BI_hotspots_dt_complete<-BI_hotspots_dt_complete[, .(mean_ddg=mean(ddg), rsasa=rsasa[1],p_value=p_value[1], p_adjusted=p_adjusted[1], median_ddg=median(ddg), BI_hotspot_class=hotspot_class[1]), by=.(lib_name, structural_alignment_pos, Uniprot_pos_ref.x)][,class:="binding_interface"]

BI_hotspots_dt_complete$lib_name<-tolower(gsub("-", "_",gsub("\n", "_", BI_hotspots_dt_complete$lib_name)))
colnames(BI_hotspots_dt_complete)[3]<-"Uniprot_pos_ref"

fwrite(BI_hotspots_dt_complete, paste0("supplementary_tables/Supplementary_table6_BI.txt"))
```









